---
title: "Electoral Cycle Effects on Parliamentary Question Behaviour"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Electoral Cycle Effects on Parliamentary Question Behaviour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  eval       = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```

> **Note:** All code chunks require a live internet connection. Fetching two
> years of written questions typically takes 10–20 seconds depending on volume
> and API rate limits. All heavy computation runs locally once the data is
> downloaded. No additional package dependencies beyond the standard
> `oiReachtas` Suggests list are required.

---

## The Dual Incentive Problem

TDs occupy a structurally awkward position. They are elected to a **national
legislature** where their formal role is to scrutinise the executive, pass
legislation, and debate policy. Yet they owe their seat to a specific
multi-seat constituency where voters expect highly personal, localised
service — fixing planning objections, chasing hospital waiting lists,
attending every parish event.

Political scientists call this the **dual mandate** (or the *clinician–
legislator* tension in comparative work). In Ireland the tension is especially
sharp because the single transferable vote in multi-seat constituencies forces
co-partisans to compete against one another: a Fianna Fáil TD in Cork South-
West must outrun not just Fine Gael but the other FF candidate on the same
ticket.

Parliamentary questions are the visible, public record of how TDs navigate
this tension. A question to the Minister for Housing about a specific estate in
Mullingar is simultaneously a national-legislature act *and* a constituency
service signal broadcast to local media. The question **title** (`show_as`)
that appears in the Oireachtas records is exactly what appears in the
candidate's next leaflet.

This vignette tests three hypotheses about how electoral proximity reshapes
that behaviour:

| # | Hypothesis | What we expect to see |
|---|---|---|
| (a) | **Volume spike** | Question counts rise sharply in the weeks before polling day |
| (b) | **Topic shift** | Housing, welfare, and local-services topics displace abstract national ones as election nears |
| (c) | **Government loyalty** | Members of government parties ask fewer critical questions of their own ministers regardless of proximity |

We operationalise "electoral proximity" as **weeks to the nearest general
election**, negative before polling day and positive after. We examine the
two most recent Dáil elections: **8 February 2020** (33rd Dáil) and
**29 November 2024** (34th Dáil).

---

## 1  Setup

```{r libraries}
library(oiReachtas)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(ggrepel)
library(scales)
library(tidytext)
```

---

## 2  Election Date Reference Table

We hardcode the two most recent Irish general elections and write a helper
that, given a vector of question dates, returns the signed distance in weeks
to the nearest election.

```{r election-dates}
elections <- tibble(
  election_name = c("2020 General Election (33rd Dáil)",
                    "2024 General Election (34th Dáil)"),
  election_date = as.Date(c("2020-02-08", "2024-11-29"))
)

# ── helper: signed weeks to nearest election ────────────────────────────────
# Negative  = weeks BEFORE the election
# Positive  = weeks AFTER  the election
# The function finds whichever election is closest to each date and returns
# that signed distance.
weeks_to_nearest_election <- function(dates, election_tbl = elections) {
  dates <- as.Date(dates)
  # Matrix: rows = question dates, cols = elections
  dist_mat <- map(election_tbl$election_date, ~ as.numeric(dates - .x) / 7)
  dist_mat <- do.call(cbind, dist_mat)   # n_dates x n_elections

  # For each date pick the election that minimises |distance|
  nearest_idx <- apply(abs(dist_mat), 1, which.min)
  signed_dist <- dist_mat[cbind(seq_along(nearest_idx), nearest_idx)]
  round(signed_dist, 2)
}
```

---

## 3  Fetch the Question Corpus

We pull **two years of written questions** spanning a window that brackets
both the pre-2020 and pre-2024 run-ups. We then attach member metadata (party,
constituency) and compute electoral proximity.

```{r fetch-questions}
# ── written questions: Jan 2019 – Dec 2024 ──────────────────────────────────
# Fetch in two blocks to stay within API page limits
qs_2019_2021 <- get_questions(
  chamber       = "dail",
  date_start    = "2019-01-01",
  date_end      = "2021-12-31",
  question_type = "written",
  limit         = 100,
  all_pages     = TRUE
)

qs_2022_2024 <- get_questions(
  chamber       = "dail",
  date_start    = "2022-01-01",
  date_end      = "2024-12-31",
  question_type = "written",
  limit         = 100,
  all_pages     = TRUE
)

qs_raw <- bind_rows(qs_2019_2021, qs_2022_2024)

# ── member roster ────────────────────────────────────────────────────────────
members <- get_members(chamber = "dail", limit = 300) |>
  select(member_name, party = party_code, constituency)

# ── join and compute electoral proximity ────────────────────────────────────
qs <- qs_raw |>
  mutate(date = as.Date(date)) |>
  left_join(members, by = "member_name") |>
  mutate(
    weeks_to_elec  = weeks_to_nearest_election(date),
    # Bin into 4-week periods for area charts (-26 to +26 covers ±6 months)
    week_bin       = floor(weeks_to_elec / 4) * 4,
    # Identify which election cycle each question belongs to
    election_cycle = if_else(date < as.Date("2022-01-01"),
                             "Pre-2020 election",
                             "Pre-2024 election")
  ) |>
  filter(abs(weeks_to_elec) <= 26)  # ± 6-month window

glimpse(qs)
```

---

## 4  Hypothesis (a): Question Volume Spike Before Elections

We aggregate weekly question counts and plot them against weeks-to-election.
The vertical line marks polling day (week 0). A positive slope approaching
week 0 would support the volume-spike hypothesis.

```{r volume-by-week, fig.height=5}
weekly_vol <- qs |>
  mutate(week_int = floor(weeks_to_elec)) |>
  count(week_int, name = "n_questions")

ggplot(weekly_vol, aes(x = week_int, y = n_questions)) +
  geom_col(fill = "#2c7bb6", alpha = 0.65, width = 0.85) +
  geom_smooth(aes(group = 1),
              method  = "loess",
              span    = 0.35,
              colour  = "#d7191c",
              linewidth = 1.1,
              se      = TRUE,
              fill    = "#d7191c",
              alpha   = 0.12) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "black",
             linewidth = 0.8) +
  annotate("text", x = 1, y = max(weekly_vol$n_questions, na.rm = TRUE) * 0.95,
           label = "Polling day", hjust = 0, size = 3.2, colour = "black") +
  scale_x_continuous(
    breaks = seq(-26, 26, by = 4),
    labels = function(x) paste0(x, "w")
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Written Question Volume Relative to Polling Day",
    subtitle = "Weekly counts pooled across 2020 and 2024 Irish general elections",
    x        = "Weeks to nearest election (negative = before)",
    y        = "Number of written questions",
    caption  = "Source: Oireachtas Open Data API via oiReachtas"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

---

## 5  Hypothesis (b): Topic Composition Shift Near Elections

We reuse the ten-domain keyword lookup from the member-profiling vignette and
track the **share of domain-tagged questions** in each 4-week bin. A rising
share for housing, social welfare, or local services as week 0 approaches
would support the topic-shift hypothesis.

```{r domain-keywords}
domains <- list(
  "Housing & Planning"    = c("housing", "rent", "home", "tenant", "landlord",
                               "accommodation", "planning", "homeless", "eviction",
                               "hap", "hal", "rents", "residential", "development"),
  "Health & HSE"          = c("health", "hospital", "hse", "waiting", "gp",
                               "mental", "nursing", "cancer", "ambulance", "pharmacy",
                               "disability", "maternity", "beds", "staffing"),
  "Education"             = c("school", "teacher", "education", "college",
                               "university", "student", "special", "sen", "curriculum",
                               "principal", "snc", "sna", "class", "pupil"),
  "Transport"             = c("road", "bus", "train", "rail", "cycle", "traffic",
                               "transport", "motorway", "greenway", "brt", "luas",
                               "dart", "airport", "port"),
  "Environment & Climate" = c("climate", "carbon", "emission", "renewable",
                               "energy", "wind", "solar", "waste", "water", "flood",
                               "biodiversity", "nature", "offshore", "fossil"),
  "Social Welfare"        = c("welfare", "pension", "disability", "carer",
                               "benefit", "payment", "allowance", "dsp", "social",
                               "lone", "jobseeker", "community", "poverty"),
  "Justice & Crime"       = c("garda", "crime", "court", "prison", "legal",
                               "justice", "solicitor", "judge", "offence", "victim",
                               "drugs", "trafficking", "violence", "fraud"),
  "Agriculture & Rural"   = c("farmer", "farm", "agri", "dairy", "beef", "rural",
                               "fishing", "forestry", "land", "mart", "tillage",
                               "dafm", "animal", "felling"),
  "Enterprise & Economy"  = c("enterprise", "business", "sme", "employment",
                               "job", "tax", "vat", "trade", "investment", "startup",
                               "export", "ibec", "isme", "commercial"),
  "Foreign & Defence"     = c("foreign", "ukraine", "conflict", "israel",
                               "palestine", "nato", "defence", "international",
                               "humanitarian", "migration", "asylum", "refugee",
                               "embassy", "united")
)

domain_lookup <- imap_dfr(domains, ~ tibble(word = .x, domain = .y))
```

```{r topic-shift}
# Tokenise show_as titles
tidy_qs <- qs |>
  select(question_id, date, week_bin, show_as) |>
  unnest_tokens(word, show_as) |>
  anti_join(stop_words, by = "word")

# Assign domain tags
domain_by_bin <- tidy_qs |>
  inner_join(domain_lookup, by = "word") |>
  count(week_bin, domain, name = "n_domain") |>
  group_by(week_bin) |>
  mutate(share = n_domain / sum(n_domain)) |>
  ungroup()

# Highlight domains of interest for electoral salience
salient_domains <- c("Housing & Planning", "Social Welfare",
                     "Health & HSE", "Transport")

domain_plot_data <- domain_by_bin |>
  mutate(domain_grp = if_else(domain %in% salient_domains,
                              domain, "Other domains"))

ggplot(domain_plot_data,
       aes(x = week_bin, y = share, fill = domain_grp)) +
  geom_area(position = "stack", alpha = 0.82) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "black",
             linewidth = 0.9) +
  scale_x_continuous(
    breaks = seq(-24, 24, by = 4),
    labels = function(x) paste0(x, "w")
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(
    values = c(
      "Housing & Planning"    = "#2c7bb6",
      "Social Welfare"        = "#abd9e9",
      "Health & HSE"          = "#fdae61",
      "Transport"             = "#d7191c",
      "Other domains"         = "#cccccc"
    )
  ) +
  labs(
    title    = "Policy Domain Composition of Written Questions by Electoral Proximity",
    subtitle = "4-week bins; stacked area = share of domain-tagged questions",
    x        = "Weeks to nearest election (negative = before)",
    y        = "Share of domain-tagged questions",
    fill     = "Policy domain",
    caption  = "Source: Oireachtas Open Data API via oiReachtas"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 2))
```

---

## 6  Hypothesis (c): Government vs Opposition Question Behaviour

Members of the governing coalition face a structural disincentive to ask
critical questions of their own ministers — doing so embarrasses colleagues,
undermines collective responsibility, and risks career consequences. We test
whether this suppression effect varies with electoral proximity.

For the **33rd Dáil** the coalition was Fianna Fáil (FF), Fine Gael (FG), and
the Green Party (GP). For the **34th Dáil** coalition formation was ongoing at
the end of the data window; we tentatively assign FF and FG to government and
treat Sinn Féin (SF) as opposition.

```{r govt-opposition}
# Government party assignments per Dáil
govt_parties_33rd <- c("FF", "FG", "GP")    # FF/FG/GP coalition
govt_parties_34th <- c("FF", "FG")           # provisional; SF in opposition

# Classify each question by govt/opposition status at time of asking
qs_gov <- qs |>
  mutate(
    dail_period = if_else(date < as.Date("2020-06-27"),
                          "33rd Dáil", "34th Dáil"),
    govt_member = case_when(
      dail_period == "33rd Dáil" & party %in% govt_parties_33rd ~ TRUE,
      dail_period == "34th Dáil" & party %in% govt_parties_34th ~ TRUE,
      TRUE                                                        ~ FALSE
    ),
    member_status = if_else(govt_member, "Government", "Opposition")
  ) |>
  filter(!is.na(party))

# Weekly question volume by status
weekly_by_status <- qs_gov |>
  mutate(week_int = floor(weeks_to_elec)) |>
  count(week_int, member_status, name = "n_questions")

ggplot(weekly_by_status,
       aes(x = week_int, y = n_questions,
           colour = member_status, fill = member_status)) +
  geom_col(position = "dodge", alpha = 0.70, width = 0.75) +
  geom_smooth(method   = "loess",
              span     = 0.4,
              se       = FALSE,
              linewidth = 1.1) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "black",
             linewidth = 0.8) +
  scale_colour_manual(values = c("Government" = "#1a9641",
                                 "Opposition" = "#d7191c")) +
  scale_fill_manual(values   = c("Government" = "#1a9641",
                                 "Opposition" = "#d7191c")) +
  scale_x_continuous(
    breaks = seq(-26, 26, by = 4),
    labels = function(x) paste0(x, "w")
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Written Question Volume: Government vs Opposition Members",
    subtitle = "Weekly counts pooled across 2020 and 2024 election windows",
    x        = "Weeks to nearest election (negative = before)",
    y        = "Number of written questions",
    colour   = "Member status",
    fill     = "Member status",
    caption  = "Source: Oireachtas Open Data API via oiReachtas"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid.minor = element_blank()
  )
```

---

## 7  Member-Level Electoral Responsiveness

Not all TDs respond equally to electoral proximity. We define an
**electoral responsiveness ratio** (ERR) for each member as:

$$\text{ERR} = \frac{\text{questions in the 8 weeks before election}}
                    {\text{questions in the 8 weeks earlier (weeks -16 to -8)}}$$

Members with ERR > 1 ramped up activity as the election approached; those
with ERR ≈ 1 maintained steady effort; those with ERR < 1 wound down early.
The chart surfaces the 20 most electorally responsive members.

```{r err-by-member}
# Window definitions (signed weeks to nearest election)
pre_election_window <- c(-8, 0)      # final 8 weeks before polling
baseline_window     <- c(-16, -8)    # 8 weeks earlier

err_data <- qs |>
  filter(!is.na(party)) |>
  mutate(
    period = case_when(
      weeks_to_elec >= baseline_window[1]     &
        weeks_to_elec <  baseline_window[2]   ~ "baseline",
      weeks_to_elec >= pre_election_window[1] &
        weeks_to_elec <  pre_election_window[2] ~ "pre_election",
      TRUE                                    ~ NA_character_
    )
  ) |>
  filter(!is.na(period)) |>
  count(member_name, party, period, name = "n_questions") |>
  pivot_wider(names_from = period, values_from = n_questions,
              values_fill = 0) |>
  # Require at least 5 questions in the baseline to avoid noise
  filter(baseline >= 5) |>
  mutate(err = pre_election / baseline) |>
  arrange(desc(err))

# Top 20 most electorally responsive members
top20_err <- slice_head(err_data, n = 20) |>
  mutate(member_name = fct_reorder(member_name, err))

# Party colour palette (standard Irish political colours)
party_colours <- c(
  "FF"  = "#006400",   # Fianna Fáil — green
  "FG"  = "#003DA5",   # Fine Gael — blue
  "SF"  = "#326760",   # Sinn Féin — dark teal
  "GP"  = "#99C140",   # Green Party — lime green
  "Lab" = "#CC0000",   # Labour — red
  "SD"  = "#7B2D8B",   # Social Democrats — purple
  "PBP" = "#8B0000",   # People Before Profit — dark red
  "Ind" = "#999999",   # Independents — grey
  "AON" = "#FF6600",   # Aontú — orange
  "NP"  = "#CCCCCC"    # No party / other
)

ggplot(top20_err,
       aes(x = err, y = member_name,
           fill = party)) +
  geom_col(alpha = 0.85) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "grey30", linewidth = 0.8) +
  scale_fill_manual(values = party_colours, na.value = "#CCCCCC") +
  scale_x_continuous(labels = number_format(accuracy = 0.1)) +
  annotate("text", x = 1.02, y = 1,
           label = "ERR = 1\n(no change)", hjust = 0,
           size = 2.9, colour = "grey30") +
  labs(
    title    = "Top 20 Most Electorally Responsive TDs",
    subtitle = "Electoral Responsiveness Ratio = questions in final 8 weeks ÷ baseline 8 weeks",
    x        = "Electoral Responsiveness Ratio (ERR)",
    y        = NULL,
    fill     = "Party",
    caption  = "Source: Oireachtas Open Data API via oiReachtas"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold"),
    legend.position  = "right",
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

---

## 8  Longitudinal Comparison: 2020 vs 2024 Run-Up

Do the patterns repeat across election cycles? We overlay the weekly question
volume trajectories for the 24 weeks before each election, normalised per
sitting TD to account for different Dáil sizes.

```{r longitudinal-comparison}
# Assign each question to its nearest election cycle
qs_cycles <- qs |>
  mutate(
    cycle = case_when(
      abs(date - as.Date("2020-02-08")) <= abs(date - as.Date("2024-11-29"))
        ~ "Pre-2020 (33rd Dáil)",
      TRUE
        ~ "Pre-2024 (34th Dáil)"
    )
  ) |>
  # Keep only the pre-election window for the comparison
  filter(weeks_to_elec >= -24 & weeks_to_elec < 0)

# Approximate TD counts per cycle for normalisation
cycle_td_counts <- tibble(
  cycle    = c("Pre-2020 (33rd Dáil)", "Pre-2024 (34th Dáil)"),
  n_tds    = c(160L, 174L)           # approx. seats at dissolution
)

longitudinal_vol <- qs_cycles |>
  mutate(week_int = floor(weeks_to_elec)) |>
  count(cycle, week_int, name = "n_questions") |>
  left_join(cycle_td_counts, by = "cycle") |>
  mutate(q_per_td = n_questions / n_tds)

ggplot(longitudinal_vol,
       aes(x = week_int, y = q_per_td,
           colour = cycle, group = cycle)) +
  geom_line(linewidth = 1.0, alpha = 0.85) +
  geom_point(size = 1.8, alpha = 0.7) +
  geom_smooth(method   = "loess",
              span     = 0.45,
              se       = FALSE,
              linewidth = 0.5,
              linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "black", linewidth = 0.8) +
  scale_colour_manual(values = c("Pre-2020 (33rd Dáil)" = "#2c7bb6",
                                 "Pre-2024 (34th Dáil)" = "#d7191c")) +
  scale_x_continuous(
    breaks = seq(-24, 0, by = 4),
    labels = function(x) paste0(x, "w")
  ) +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) +
  labs(
    title    = "Pre-Election Question Surge: 2020 vs 2024",
    subtitle = "Weekly written questions per sitting TD, final 24 weeks before polling day",
    x        = "Weeks to election",
    y        = "Written questions per TD",
    colour   = "Election cycle",
    caption  = "Source: Oireachtas Open Data API via oiReachtas"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold"),
    legend.position  = "top",
    panel.grid.minor = element_blank()
  )
```

---

## 9  Summary of Findings

```{r findings-table, eval=FALSE}
# Compute headline statistics for a findings summary
findings <- tribble(
  ~hypothesis,     ~metric,                            ~expected_sign,
  "(a) Volume spike",
    "Pearson r: weekly questions ~ weeks_to_elec (pre-election only)",
    "Negative (more Qs as election nears)",
  "(b) Topic shift",
    "Change in housing+welfare share: bin[-8,0] vs bin[-24,-16]",
    "Positive (higher share near election)",
  "(c) Government loyalty",
    "Govt Q volume / Opposition Q volume ratio near election",
    "< 1 (govt members ask fewer Qs)"
)

# Compute (a)
cor_a <- qs |>
  filter(weeks_to_elec >= -26, weeks_to_elec < 0) |>
  mutate(week_int = floor(weeks_to_elec)) |>
  count(week_int) |>
  summarise(r = cor(week_int, n, method = "pearson")) |>
  pull(r)

# Compute (b)
share_shift <- domain_by_bin |>
  filter(domain %in% c("Housing & Planning", "Social Welfare")) |>
  group_by(week_bin) |>
  summarise(combined_share = sum(share), .groups = "drop") |>
  summarise(
    near_election = mean(combined_share[week_bin >= -8 & week_bin < 0]),
    early_window  = mean(combined_share[week_bin >= -24 & week_bin < -16])
  ) |>
  mutate(shift = near_election - early_window)

# Compute (c)
vol_ratio <- weekly_by_status |>
  filter(week_int >= -8, week_int < 0) |>
  group_by(member_status) |>
  summarise(total = sum(n_questions), .groups = "drop") |>
  pivot_wider(names_from = member_status, values_from = total) |>
  mutate(ratio = Government / Opposition)

cat("=== Hypothesis (a) — Volume spike ===\n")
cat("  Pearson r (weekly count ~ weeks to election, pre-election):",
    round(cor_a, 3), "\n")
cat("  Interpretation: r < 0 supports H(a)\n\n")

cat("=== Hypothesis (b) — Topic shift ===\n")
cat("  Housing+Welfare share change (early vs near-election):",
    round(share_shift$shift * 100, 1), "pp\n")
cat("  Interpretation: positive shift supports H(b)\n\n")

cat("=== Hypothesis (c) — Government loyalty ===\n")
cat("  Govt/Opposition question ratio (final 8 weeks):",
    round(vol_ratio$ratio, 3), "\n")
cat("  Interpretation: ratio < 1 supports H(c)\n")
```

---

## 10  Caveats and Limitations

Several features of the parliamentary calendar and question system complicate
a clean causal interpretation of these patterns.

**Parliamentary recess.** The Dáil does not sit continuously. The summer
recess (roughly July–September) and Christmas recess (roughly late December–
early January) produce zero-question weeks that have nothing to do with
electoral incentives. A dissolved Dáil similarly generates no questions.
Ensure that any trend model conditions on sitting weeks, or at minimum flags
zero-activity weeks as structural rather than behavioural zeros.

**Question allocation limits.** The Oireachtas Business Committee sets limits
on the number of questions each member may table per week, which varies by
standing order period and by whether questions are oral or written. A surge in
*submitted* questions could be partially absorbed by scheduling, so observed
*answered* question counts are a censored measure of member effort.

**Strategic timing.** Not all pre-election questions are electorally motivated.
Ministers are required to answer questions tabled before dissolution even
after an election is called, so members may front-load high-priority questions
for administrative rather than electoral reasons.

**Coalition uncertainty.** Government membership changes at dissolution and
again after negotiation. The classification of members as government or
opposition is therefore uncertain in the weeks immediately before and after
polling day, particularly for parties that might enter multiple different
coalitions across the two cycles examined here.

**Selection on questions asked.** Questions visible in the API are *tabled*
questions — those accepted and scheduled. Questions withdrawn, ruled
inadmissible, or never submitted are unobservable. If electoral pressure
causes members to submit more questions simultaneously (hoping some get
scheduled), the API record understates the true surge.

**Individual vs collective incentives.** Members in safe seats or retiring
members face very different incentive gradients. Ideally the volume and topic
analyses should be stratified by seat marginality, but constituency-level
vote-share data would need to be matched from an external election results
database.

---

## Further Reading

- Gallagher, M. & Komito, L. (2010). *The constituency role of TDs.* In
  Gallagher & Coakley (eds.), *Politics in the Republic of Ireland*, 5th ed.
- Martin, S. (2011). *Electoral incentives and parliamentary activity: lessons
  from the Irish case.* Parliamentary Affairs, 64(3), 512–534.
- Saalfeld, T. (2011). *Parliamentary questions as instruments of substantive
  representation.* Journal of Legislative Studies, 17(3), 295–322.
- Oireachtas Open Data API documentation: <https://data.oireachtas.ie>

---

```{r session-info}
sessionInfo()
```
