---
title: "Focus vs. Reactivity: Identifying Specialists and Populists in the Dáil"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Focus vs. Reactivity: Identifying Specialists and Populists in the Dáil}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```

> **Note:** All code chunks require a live internet connection.
> The vignette reuses the domain lookup and TF-IDF pipeline introduced in
> *Member Profiling* — run that vignette first, or re-run sections 2–4 here
> to rebuild the required objects.

---

## The Specialist vs. Populist Question

Two TDs can both ask about housing without resembling each other at all.
One has asked about housing for four consecutive years regardless of what's
in the headlines — she built a reputation as the go-to voice on planning
reform long before it became a crisis.
The other pivoted to housing questions the week a record rental index was
published, having spent the previous quarter asking about broadband, tourism,
and data centres.

These are fundamentally different parliamentary behaviours.
Distinguishing them requires measuring **two orthogonal dimensions**:

| Dimension | High score means | Low score means |
|-----------|-----------------|-----------------|
| **Focus** | Questions concentrated in few policy areas | Questions spread across many areas |
| **Consistency** | Same topics quarter after quarter | Topics shift with news cycles |

Crossing these two dimensions produces four archetypes:

```
                HIGH CONSISTENCY
                       │
    Deep specialist    │    Steady generalist
    "Owns the brief"   │    "Reliable workhorse"
                       │
LOW FOCUS ─────────────┼───────────────── HIGH FOCUS
                       │
    Populist           │    Bandwagon rider
    "Chases headlines" │    "Latest hot take"
                       │
                LOW CONSISTENCY
```

A third dimension — **herd instinct** — measures whether a member's monthly
topic distribution tracks or diverges from the Dáil-wide topic surge.
High herd instinct is a hallmark of the populist quadrant even when that
member's raw focus score looks deceptively high.

---

## 1  Setup

```{r libraries}
library(oiReachtas)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(ggrepel)
library(scales)
library(tidytext)
```

---

## 2  Data

```{r fetch}
# ── Member roster: biographical + party/constituency ──────────────────────

# ── 12 months of written questions ──────────────────────────────────────────

fetch_quarter <- function(start, end) {
  get_questions(date_start = start, date_end = end,
                question_type = "written", limit = 100, all_pages = TRUE)
}

qs_2022_2024  <- bind_rows(
  fetch_quarter("2022-01-01", "2022-03-30"),
  fetch_quarter("2022-04-01", "2022-06-30"),
  fetch_quarter("2022-07-01", "2022-09-30"),
  fetch_quarter("2022-10-01", "2022-12-31"),
  fetch_quarter("2023-01-01", "2023-03-30"),
  fetch_quarter("2023-04-01", "2023-06-30"),
  fetch_quarter("2023-07-01", "2023-09-30"),
  fetch_quarter("2023-10-01", "2023-12-31"),
  fetch_quarter("2024-01-01", "2024-03-30"),
  fetch_quarter("2024-04-01", "2024-06-30"),
  fetch_quarter("2024-07-01", "2024-09-30"),
  fetch_quarter("2024-10-01", "2024-12-31")
)

qs_raw <- qs_2022_2024

# ── member roster: biographical + party/constituency ───────────────────────
# get_members() provides names; get_member_parties() provides affiliations
members <- get_members(chamber = "dail", house_no = 33, all_pages = TRUE) 

member_ids <- members$member_code|>unique()

party_details <- lapply(member_ids,\(x) get_member_parties(member_id = x))|>
  do.call(what= rbind)

current_party <- party_details|>
 filter(house_no == 33)|>
  group_by(member_id)|>
  filter(row_number() == 1)





qs <- qs_raw |>
  select(member_name,member_uri, show_as, department,date) |>
  left_join(current_party, by = c("member_uri" = "member_id")) |>
  filter(!is.na(member_name), !is.na(show_as), nchar(show_as) > 5)|>
  mutate(
    date    = as.Date(date),
    year    = format(date, "%Y"),
    month   = format(date, "%Y-%m"),
    quarter = paste0(format(date, "%Y"), "-Q",
                     ceiling(as.integer(format(date, "%m")) / 3))
  )

# Minimum questions threshold — members with very few questions give noisy signals
MIN_QS <- 20
active_members <- qs |> count(member_name) |> filter(n >= MIN_QS) |> pull(member_name)
qs <- filter(qs, member_name %in% active_members)

cat(nrow(qs), "questions from", length(active_members), "members\n")
```

---

## 3  Domain Scoring

We reuse the ten-domain keyword lookup from the *Member Profiling* vignette.

```{r domains}
oireachtas_stops <- tibble(word = c(
  "minister", "government", "plans", "measures", "steps", "action", "actions",
  "information", "details", "aware", "issues", "concerned", "basis", "provision",
  "provisions", "matter", "matters", "support", "supports", "service", "services",
  "scheme", "schemes", "regard", "relevant", "area", "areas", "funding", "ensure",
  "provide", "policy", "policies", "report", "request", "asked", "stated",
  "confirm", "outlined"
))

domains <- list(
  "Housing"       = c("housing", "rent", "home", "tenant", "landlord",
                      "accommodation", "planning", "homeless", "eviction",
                      "hap", "hal", "rents", "residential", "development"),
  "Health"        = c("health", "hospital", "hse", "waiting", "gp", "mental",
                      "nursing", "cancer", "ambulance", "pharmacy", "disability",
                      "maternity", "beds", "staffing"),
  "Education"     = c("school", "teacher", "education", "college", "university",
                      "student", "special", "sen", "curriculum", "sna", "pupil"),
  "Transport"     = c("road", "bus", "train", "rail", "cycle", "traffic",
                      "transport", "motorway", "greenway", "luas", "dart"),
  "Environment"   = c("climate", "carbon", "emission", "renewable", "energy",
                      "wind", "solar", "waste", "water", "flood", "biodiversity"),
  "Social Welfare"= c("welfare", "pension", "disability", "carer", "benefit",
                      "payment", "allowance", "dsp", "jobseeker", "poverty"),
  "Justice"       = c("garda", "crime", "court", "prison", "legal", "justice",
                      "solicitor", "offence", "victim", "drugs", "violence"),
  "Agriculture"   = c("farmer", "farm", "agri", "dairy", "beef", "rural",
                      "fishing", "forestry", "land", "tillage", "dafm"),
  "Enterprise"    = c("enterprise", "business", "sme", "employment", "job",
                      "tax", "vat", "trade", "investment", "startup", "export"),
  "Foreign"       = c("foreign", "ukraine", "conflict", "israel", "palestine",
                      "defence", "international", "migration", "asylum", "refugee")
)

domain_lookup <- imap_dfr(domains, ~ tibble(word = .x, domain = .y))
K_DOMAINS <- length(domains)  # 10

# Tidy question corpus with domain tags
tidy_qs <- qs |>
  unnest_tokens(word, show_as) |>
  anti_join(stop_words,       by = "word") |>
  anti_join(oireachtas_stops, by = "word") |>
  filter(nchar(word) > 2, !grepl("^[0-9]+$", word))

domain_words <- tidy_qs |>
  inner_join(domain_lookup, by = "word")
```

---

## 4  Focus Score — Herfindahl-Hirschman Index

The **HHI** (borrowed from market-concentration economics) measures how
unevenly a member's questions are distributed across domains.

* HHI = Σ (share_i)²
* Ranges from **1/k ≈ 0.1** (perfectly uniform) to **1.0** (single domain)
* We normalise to 0–1: `(HHI − 1/k) / (1 − 1/k)`

A score of 1 means every question is in the same domain.
A score of 0 means questions are uniformly spread across all ten.

```{r hhi}
member_domain_shares <- domain_words |>
  count(member_name, domain, name = "n") |>
  group_by(member_name) |>
  # Fill domains with 0 questions
  complete(domain = names(domains), fill = list(n = 0)) |>
  mutate(share = n / sum(n)) |>
  ungroup()

focus_scores <- member_domain_shares |>
  group_by(member_name) |>
  summarise(
    hhi       = sum(share^2),
    hhi_norm  = (hhi - 1/K_DOMAINS) / (1 - 1/K_DOMAINS),
    top_domain = domain[which.max(share)],
    top_share  = max(share),
    .groups    = "drop"
  )

# Distribution
ggplot(focus_scores, aes(hhi_norm)) +
  geom_histogram(bins = 30, fill = "#5C6BC0", colour = "white") +
  geom_vline(xintercept = median(focus_scores$hhi_norm),
             linetype = "dashed", colour = "grey40") +
  annotate("text", x = median(focus_scores$hhi_norm) + 0.01,
           y = Inf, vjust = 1.5, hjust = 0,
           label = "median", size = 3, colour = "grey30") +
  labs(
    title    = "Distribution of member focus scores (normalised HHI)",
    subtitle = "0 = questions spread uniformly · 1 = all questions in one domain",
    x        = "Focus score (normalised HHI)", y = "Members"
  ) +
  theme_minimal()
```

```{r focus-bar}

members_clean <- qs|>
  group_by(member_uri)|>
  filter(row_number() == 1)|>
  select(member_name,member_uri,party_name,constituency)|>
  mutate(
    party_code = case_when(
      stringr::str_detect(party_name|>tolower(), "fianna|FF") ~ "FF",
      stringr::str_detect(party_name|>tolower(), "fine gael|FG") ~ "FG",
      stringr::str_detect(party_name|>tolower(), "sinn féin|SF") ~ "SF",
      stringr::str_detect(party_name|>tolower(), "labour|Lab") ~ "Lab",
      stringr::str_detect(party_name|>tolower(), "green|GP") ~ "GP",
      stringr::str_detect(party_name|>tolower(), "social dem|SD") ~ "SD",
      stringr::str_detect(party_name|>tolower(), "people before|PBP") ~ "PBP",
      TRUE ~ "Ind"
    )
  )


focus_scores |>
  left_join(members_clean, by = "member_name") |>
  slice_max(hhi_norm, n = 30) |>
  mutate(member_name = reorder(member_name, hhi_norm)) |>
  ggplot(aes(hhi_norm, member_name, fill = party_code)) +
  geom_col() +
  geom_text(aes(label = top_domain), hjust = -0.05, size = 2.8, colour = "grey30") +
  scale_fill_manual(
    values = c("FF" = "#66BB6A", "FG" = "#42A5F5", "SF" = "#4CAF50",
               "Lab" = "#EF5350", "GP" = "#26A69A", "SD" = "#AB47BC",
               "PBP" = "#FF7043", "Ind" = "#78909C"),
    na.value = "#BDBDBD"
  ) +
  scale_x_continuous(limits = c(0, 1.05), labels = label_percent(accuracy = 1)) +
  labs(
    title    = "Top-30 most focused members",
    subtitle = "Bar shows normalised HHI; label = dominant policy domain",
    x        = "Focus score", y = NULL, fill = "Party",
    caption  = "Source: Oireachtas Open Data API — written questions 2023–2024"
  ) +
  theme_minimal(base_size = 11)
```

---

## 5  Consistency Score — Temporal Stability

We split each member's question corpus into quarters and compute the
**domain vector** (share per domain) for each quarter.  The consistency
score is the **mean cosine similarity between consecutive quarters'
domain vectors**.

* Score near **1.0** → same topic distribution every quarter → specialist
* Score near **0.0** → topic distribution changes radically → reactive

```{r consistency}
# Per-member per-quarter domain shares
quarterly_domains <- domain_words |>
  count(member_name, quarter, domain, name = "n") |>
  group_by(member_name, quarter) |>
  complete(domain = names(domains), fill = list(n = 0)) |>
  mutate(share = n / (sum(n) + 1e-9)) |>
  ungroup()

# Pivot to wide so each row is a member-quarter vector
quarterly_wide <- quarterly_domains |>
  select(member_name, quarter, domain, share) |>
  pivot_wider(names_from = domain, values_from = share, values_fill = 0)

# Cosine similarity between two numeric vectors
cos_sim <- function(a, b) {
  sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2)) + 1e-10)
}

# For each member, compute mean cosine similarity across consecutive quarters
quarters_ordered <- sort(unique(quarterly_wide$quarter))

consistency_scores <- quarterly_wide |>
  group_by(member_name) |>
  group_map(~ {
    # .x contains the data for one member
    df <- arrange(.x, quarter)
    
    if (nrow(df) < 2) return(tibble(consistency = NA_real_))
    
    # Select ONLY numeric columns (the domain shares)
    vecs <- df |> 
      select(where(is.numeric)) |> 
      as.matrix()
    
    # Compute consecutive similarities
    sims <- map_dbl(seq_len(nrow(vecs) - 1),
                    ~ cos_sim(vecs[.x, ], vecs[.x + 1, ]))
    
    tibble(consistency = mean(sims, na.rm = TRUE))
  }, .keep = TRUE) |>
  bind_rows() |>
  mutate(member_name = sort(unique(quarterly_wide$member_name)))
```

---

## 6  Department Diversity

The number of **distinct government departments** a member directs questions
to is a coarse but intuitive measure of spread.  Specialists interrogate
1–2 departments relentlessly; generalists (and populists) spray across many.

```{r dept-diversity}
dept_diversity <- qs |>
  group_by(member_name) |>
  summarise(
    n_depts        = n_distinct(department, na.rm = TRUE),
    n_qs           = n(),
    depts_per_100q = n_depts / n_qs * 100,
    .groups        = "drop"
  )

ggplot(dept_diversity, aes(n_qs, n_depts)) +
  geom_point(alpha = 0.55, colour = "#5C6BC0") +
  geom_text_repel(
    data = slice_max(dept_diversity, n_depts, n = 15),
    aes(label = member_name), size = 2.5, max.overlaps = 20
  ) +
  scale_x_log10(labels = label_comma()) +
  labs(
    title    = "Questions filed vs. departments addressed",
    subtitle = "Members top-right spread questions across many departments",
    x        = "Total written questions (log scale)",
    y        = "Distinct departments questioned"
  ) +
  theme_minimal()
```

---

## 7  The Focus × Consistency Map

This is the central chart — the political equivalent of a shooting-efficiency
vs. usage-rate scatter in basketball analytics.

```{r focus-consistency-map}
# Combine all three scores
style_scores <- focus_scores |>
  select(member_name, focus = hhi_norm, top_domain) |>
  inner_join(consistency_scores, by = "member_name") |>
  inner_join(select(dept_diversity, member_name, depts_per_100q), by = "member_name") |>
  left_join(members_clean, by = "member_name") |>
  filter(!is.na(consistency))

# Medians — used for quadrant lines
med_focus <- median(style_scores$focus)
med_cons  <- median(style_scores$consistency)

# Quadrant labels
quadrant_labels <- tibble(
  x     = c(med_focus / 2,      (1 + med_focus) / 2,
            med_focus / 2,      (1 + med_focus) / 2),
  y     = c(med_cons + 0.05,    med_cons + 0.05,
            med_cons - 0.05,    med_cons - 0.05),
  label = c("Steady generalist\n(broad, consistent)",
            "Deep specialist\n(focused, consistent)",
            "Populist\n(broad, reactive)",
            "Bandwagon rider or Agile Campaigner\n(focused, reactive)")
)

party_colours <- c(
  "FF" = "#66BB6A", "FG" = "#42A5F5", "SF" = "#4CAF50",
  "Lab" = "#EF5350", "GP" = "#26A69A", "SD" = "#AB47BC",
  "PBP" = "#FF7043", "Ind" = "#78909C"
)

ggplot(style_scores, aes(focus, consistency, colour = party_code,
                          label = member_name)) +
  # Quadrant lines
  geom_vline(xintercept = med_focus, linetype = "dashed",
             colour = "grey60", linewidth = 0.5) +
  geom_hline(yintercept = med_cons,  linetype = "dashed",
             colour = "grey60", linewidth = 0.5) +
  # Quadrant labels
  geom_text(data = quadrant_labels,
            aes(x = x, y = y, label = label),
            colour = "grey45", size = 3, fontface = "italic",
            lineheight = 0.85, inherit.aes = FALSE) +
  # Members
  geom_point(size = 2.5, alpha = 0.85) +
  geom_text_repel(size = 2.4, max.overlaps = 18, segment.colour = "grey75") +
  scale_colour_manual(values = party_colours, na.value = "#BDBDBD") +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  scale_y_continuous(labels = label_number(accuracy = 0.01)) +
  labs(
    title    = "Parliamentary style: focus vs. consistency",
    subtitle = paste0(
      "Focus = HHI of domain shares (higher = more concentrated)  ·  ",
      "Consistency = mean consecutive-quarter cosine similarity (higher = same topics)"),
    x        = "Focus score (normalised HHI)",
    y        = "Consistency score (mean consecutive-quarter cosine sim)",
    colour   = "Party",
    caption  = "Source: Oireachtas Open Data API — written questions 2023–2024"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right", plot.subtitle = element_text(size = 8.5))
```

---

## 8  Herd Instinct — Does the Member Track Dáil-Wide Topic Surges?

For each month we compute the **Dáil-wide topic distribution** — what
fraction of all questions across all members touched each domain.
We then compare each member's monthly topic distribution to this
"parliamentary zeitgeist".

A high **herd instinct score** (mean cosine similarity to the Dáil-wide
monthly vector) means the member closely mirrors what the whole chamber is
asking about — an independent signal of reactive, news-driven behaviour.

```{r herd-instinct}
# ── Dáil-wide monthly domain fractions (the "zeitgeist") ────────────────────
dail_monthly <- domain_words |>
  count(month, domain, name = "n") |>
  group_by(month) |>
  complete(domain = names(domains), fill = list(n = 0)) |>
  mutate(dail_share = n / (sum(n) + 1e-9)) |>
  ungroup() |>
  select(month, domain, dail_share)

# ── Per-member monthly domain fractions ─────────────────────────────────────
member_monthly <- domain_words |>
  count(member_name, month, domain, name = "n") |>
  group_by(member_name, month) |>
  complete(domain = names(domains), fill = list(n = 0)) |>
  mutate(member_share = n / (sum(n) + 1e-9)) |>
  ungroup()

# ── Cosine similarity of member vector vs. Dáil-wide vector, per month ──────
herd_scores <- member_monthly |>
  left_join(dail_monthly, by = c("month", "domain")) |>
  group_by(member_name, month) |>
  summarise(
    monthly_cos = cos_sim(member_share, dail_share),
    .groups = "drop"
  ) |>
  group_by(member_name) |>
  summarise(
    herd_instinct = mean(monthly_cos, na.rm = TRUE),
    n_active_months = n(),
    .groups = "drop"
  ) |>
  filter(n_active_months >= 4)   # need enough months for a meaningful average

# Distribution
ggplot(herd_scores, aes(herd_instinct)) +
  geom_histogram(bins = 25, fill = "#EF5350", colour = "white") +
  geom_vline(xintercept = median(herd_scores$herd_instinct),
             linetype = "dashed", colour = "grey40") +
  labs(
    title    = "Herd instinct distribution",
    subtitle = paste0(
      "High score = member's monthly topics closely track Dáil-wide topic surges\n",
      "Low score  = idiosyncratic focus, independent of what others are asking about"),
    x = "Herd instinct score (mean cosine sim to monthly Dáil vector)",
    y = "Members"
  ) +
  theme_minimal()
```

```{r herd-vs-focus}
# Does herd instinct correlate with (or diverge from) the focus score?
style_scores |>
  inner_join(herd_scores, by = "member_name") |>
  ggplot(aes(focus, herd_instinct, colour = party_code, label = member_name)) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_text_repel(size = 2.4, max.overlaps = 15, segment.colour = "grey75") +
  geom_smooth(method = "lm", se = TRUE, colour = "grey40",
              linetype = "dashed", linewidth = 0.8) +
  scale_colour_manual(values = party_colours, na.value = "#BDBDBD") +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title    = "Focus score vs. herd instinct",
    subtitle = paste0(
      "Top-right = focused AND tracks Dáil trends (bandwagon rider on a single issue)\n",
      "Bottom-left = unfocused AND independent (idiosyncratic generalist)"),
    x      = "Focus score (normalised HHI)",
    y      = "Herd instinct",
    colour = "Party"
  ) +
  theme_minimal(base_size = 11)
```

---

## 9  Pioneer Index — Who Sets the Agenda?

Beyond tracking averages, we want to know who **raises topics before they
become newsworthy** vs. who **piles on after the surge**.

### Method

1. For each domain, compute the Dáil-wide **monthly question rate** (questions
   in that domain ÷ all questions that month).
2. Detect **breakout months** — months where a domain's rate exceeds
   1.5× its trailing 3-month average (a meaningful surge, not noise).
3. For each breakout event, check which members were already asking above
   their personal average in that domain **in the month before** the surge
   (pioneers) vs. **in the month of or after** the surge (followers).
4. **Pioneer index** = (pioneer events − follower events) / total events.
   Ranges from −1 (pure follower) to +1 (pure pioneer).

```{r pioneer}
# ── Dáil-wide domain rate by month ──────────────────────────────────────────
dail_rate <- domain_words |>
  count(month, domain, name = "domain_n") |>
  left_join(count(qs, month, name = "total_n"), by = "month") |>
  mutate(rate = domain_n / total_n) |>
  arrange(domain, month) |>
  group_by(domain) |>
  mutate(
    trailing_avg = slider::slide_mean(rate, before = 3, after = 0,
                                      complete = FALSE),
    is_breakout  = rate > 1.5 * lag(trailing_avg, 1)
  ) |>
  ungroup()

# ── Member monthly domain rate ───────────────────────────────────────────────
member_rate <- domain_words |>
  count(member_name, month, domain, name = "member_n") |>
  left_join(count(qs, member_name, month, name = "member_total"),
            by = c("member_name", "month")) |>
  mutate(member_rate = member_n / member_total) |>
  group_by(member_name, domain) |>
  mutate(member_avg = mean(member_rate, na.rm = TRUE)) |>
  ungroup()

# ── Classify each member at each breakout event ──────────────────────────────
breakout_events <- dail_rate |>
  filter(is_breakout) |>
  select(domain, breakout_month = month)

# For each breakout, look at member activity in the month prior
classify_at_breakout <- breakout_events |>
  left_join(member_rate, by = "domain") |>
  # The month prior to the breakout
  mutate(
    pre_month = format(as.Date(paste0(breakout_month, "-01")) - 31, "%Y-%m")
  ) |>
  filter(month == pre_month) |>
  mutate(
    was_active_pre = member_rate > member_avg,   # above their personal average
    classification = if_else(was_active_pre, "pioneer", "follower")
  )

pioneer_index <- classify_at_breakout |>
  group_by(member_name) |>
  summarise(
    n_pioneer  = sum(classification == "pioneer"),
    n_follower = sum(classification == "follower"),
    n_events   = n(),
    pioneer_index = (n_pioneer - n_follower) / (n_events + 1e-9),
    .groups = "drop"
  ) |>
  filter(n_events >= 3)   # need enough events to be meaningful
```

> **Note:** The `slider` package is required for `slider::slide_mean()`.
> Install with `install.packages("slider")`.
> As an alternative, replace with `zoo::rollmeanr()` or a manual 3-month
> lag: `(lag(rate,1) + lag(rate,2) + lag(rate,3)) / 3`.

```{r pioneer-plot}
pioneer_index |>
  left_join(members_clean, by = "member_name") |>
  slice_max(abs(pioneer_index), n = 40) |>
  mutate(
    member_name = reorder(member_name, pioneer_index),
    direction   = if_else(pioneer_index >= 0, "Pioneer", "Follower")
  ) |>
  ggplot(aes(pioneer_index, member_name, fill = direction)) +
  geom_col() +
  geom_vline(xintercept = 0, colour = "grey40") +
  scale_fill_manual(values = c("Pioneer" = "#26A69A", "Follower" = "#EF5350")) +
  scale_x_continuous(labels = label_number(accuracy = 0.01)) +
  labs(
    title    = "Pioneer index — who sets vs. follows the agenda?",
    subtitle = paste0(
      "Positive = member was raising the topic before Dáil-wide surges\n",
      "Negative = member piled on after the topic was already trending"),
    x        = "Pioneer index  (−1 = pure follower · +1 = pure pioneer)",
    y        = NULL, fill = NULL,
    caption  = "Source: Oireachtas Open Data API — written questions 2023–2024"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

---

## 10  Combined Parliamentary Style Score

Combine all four signals into a single **style summary table** for easy
cross-comparison.

```{r combined-table}
# ── Final summary ─────────────────────────────────────────────────────────
style_summary <- style_scores |>
  inner_join(herd_scores,   by = "member_name") |>
  left_join(pioneer_index |> select(member_name, pioneer_index),
            by = "member_name") |>
  left_join(select(dept_diversity, member_name, n_depts, depts_per_100q),
            by = "member_name") |>
  left_join(members_clean,
            by = "member_name") |>
  mutate(
    style_quadrant = case_when(
      focus >= med_focus & consistency >= med_cons ~ "Deep specialist",
      focus <  med_focus & consistency >= med_cons ~ "Steady generalist",
      focus >= med_focus & consistency <  med_cons ~ "Bandwagon rider",
      TRUE                                         ~ "Populist"
    )
  ) |>
  transmute(member_name, party_code = party_code.x, constituency = constituency.x,
         focus, consistency , herd_instinct, pioneer_index,
         n_depts, top_domain, style_quadrant) |>
  arrange(desc(focus), desc(consistency))

# Print top specialists
style_summary |>
  filter(style_quadrant == "Deep specialist") |>
  select(member_name, party_code, top_domain, focus, consistency,
         herd_instinct, pioneer_index) |>
  mutate(across(where(is.double), ~ round(.x, 3))) |>
  print(n = 20)
```

### 10.1  Quadrant composition by party

```{r quadrant-party}
style_summary |>
  count(party_code, style_quadrant) |>
  group_by(party_code) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(style_quadrant, pct, fill = party_code)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = party_colours, na.value = "#BDBDBD") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "Parliamentary style quadrant composition by party",
    subtitle = "% of each party's members in each quadrant",
    x = NULL, y = "% of party members", fill = "Party"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

### 10.2  Longitudinal view — are members becoming more or less focused?

```{r longitudinal}
# Compute HHI per member per year to spot drift
annual_hhi <- domain_words |>
  count(member_name, year, domain, name = "n") |>
  group_by(member_name, year) |>
  complete(domain = names(domains), fill = list(n = 0)) |>
  mutate(share = n / (sum(n) + 1e-9)) |>
  summarise(hhi_norm = (sum(share^2) - 1/K_DOMAINS) / (1 - 1/K_DOMAINS),
            .groups = "drop")

annual_hhi |>
  inner_join(members_clean, by = "member_name") |>
  ggplot(aes(year, hhi_norm, group = member_name, colour = party_code)) +
  geom_line(alpha = 0.4, linewidth = 0.6) +
  stat_summary(aes(group = party_code), fun = mean,
               geom = "line", linewidth = 1.2) +
  stat_summary(aes(group = party_code), fun = mean,
               geom = "point", size = 2.5) +
  scale_colour_manual(values = party_colours, na.value = "#BDBDBD") +
  scale_y_continuous(labels = label_number(accuracy = 0.01)) +
  facet_wrap(~party_code, ncol = 4) +
  labs(
    title    = "Member focus score over time (by year)",
    subtitle = "Thin lines = individual members · Bold line = party average",
    x = NULL, y = "Focus score (normalised HHI)", colour = "Party"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")
```

---

## Summary

| Metric | Method | Interpretation |
|--------|--------|---------------|
| **Focus score** | Normalised HHI on domain shares | 1 = all questions in one domain |
| **Consistency** | Mean consecutive-quarter cosine similarity | 1 = same topic distribution every quarter |
| **Herd instinct** | Mean cosine sim to monthly Dáil-wide vector | High = mirrors parliamentary zeitgeist |
| **Pioneer index** | Pre- vs. post-surge activity at breakout events | Positive = sets agenda; negative = follows |
| **Style quadrant** | Focus × Consistency median split | 4 behavioural archetypes |

The four archetypes:

* **Deep specialist** — owns a brief, consistent over time, low herd instinct, often positive pioneer index
* **Steady generalist** — broad portfolio but stable; the reliable workhorse
* **Bandwagon rider** — concentrates on whichever topic is currently most salient, then pivots
* **Populist** — broad AND reactive; questions spray across every topic as each one rises in headlines

None of these labels is inherently pejorative — a responsive generalist
serving a diverse constituency may rationally adopt a populist pattern.
The value is in the transparency: constituents and journalists can see the
pattern in the data rather than relying on the member's self-description.
