---
title: "Legislative Tracking: From Introduction to Enactment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Legislative Tracking: From Introduction to Enactment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE   # all chunks require a live internet connection
)
```

> **Note:** Every code block in this vignette makes live API calls. Run them
> interactively in an R session with an internet connection. Results are cached
> for one hour by default, so re-running a script is fast.

## The Irish Legislative Process

A bill in Ireland travels through a fixed series of stages before it can
become law:

```
                    Dáil Éireann                    Seanad Éireann
 ┌──────────────────────────────────┐  ┌──────────────────────────────────┐
 │ 1st Reading  (formal intro)      │  │ 1st Reading                      │
 │ 2nd Reading  (principle debate)  │─▶│ 2nd Reading                      │
 │ Committee    (line-by-line)      │  │ Committee                        │
 │ Report Stage (amendments)        │  │ Report Stage                     │
 │ 5th Stage    (final vote)        │  │ 5th Stage                        │
 └──────────────────────────────────┘  └──────────────────────────────────┘
                                                       │
                                             Presidential Assent
                                                       │
                                                     ACT ✓
```

Government Bills are introduced by a minister; Private Members' Bills (PMBs)
are introduced by individual TDs or Senators. Only a fraction of PMBs are
ever enacted.

The Oireachtas Open Data API records the **current status** of every bill and,
for enacted legislation, the **Act number** and **year**. We can use these to
build a full-lifecycle view.

---

## 1  Setup

```{r libraries}
library(oiReachtas)
library(dplyr)
library(ggplot2)
```

---

## 2  Fetching the Legislation Corpus

### 2.1  A snapshot of the current Dáil

Bills before the 33rd Dáil (dissolved in 2025) span a representative modern
period. We fetch the complete set — the session cache means subsequent calls
are instant.

```{r fetch-all}
legs_33 <- get_legislation(chamber = '32',all_pages = TRUE)

glimpse(legs_33)
#> Rows: 412
#> Columns: 12
#> $ bill_id         <chr> "/ie/oireachtas/bill/2020/2", …
#> $ bill_no         <chr> "2", "7", …
#> $ bill_year       <chr> "2020", "2020", …
#> $ bill_type       <chr> "Government", "Private Members'", …
#> $ bill_status     <chr> "Enacted", "Lapsed", …
#> $ title           <chr> "Criminal Justice (Money Laundering …", …
#> $ title_ga        <chr> NA, NA, …
#> $ chamber         <chr> "Dáil Éireann", "Seanad Éireann", …
#> $ sponsor_name    <chr> "Minister for Justice", "Rónán Mullen", …
#> $ date_introduced <chr> "2020-02-06", …
#> $ act_no          <chr> "9", NA, …
#> $ act_year        <chr> "2021", NA, …
```

---

## 3  The Status Pipeline

Most bills never reach the statute book. Visualising the pipeline makes this
stark.

```{r status-order}
# Define a meaningful reading order for the y-axis
status_levels <- c(
  "Current", "Enacted", "Defeated", "Withdrawn", "Lapsed"
)

status_counts <- legs_33 |>
  count(bill_status, name = "n_bills") |>
  mutate(
    bill_status = factor(bill_status, levels = rev(status_levels)),
    label       = as.character(n_bills)
  ) |>
  arrange(bill_status)
```

```{r plot-pipeline, fig.width=7, fig.height=4}
ggplot(status_counts, aes(x = n_bills, y = bill_status, fill = bill_status)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = label), hjust = -0.2, size = 3.5) +
  scale_fill_manual(values = c(
    "Enacted"   = "#2a9d8f",
    "Current"   = "#457b9d",
    "Defeated"  = "#e76f51",
    "Withdrawn" = "#f4a261",
    "Lapsed"    = "#bdbdbd"
  )) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "33rd Dáil: Bills by Final Status",
    subtitle = "Most private members' bills lapse or are withdrawn",
    x        = "Number of bills",
    y        = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

---

## 4  Government vs Private Members' Bills

```{r bill-type-status}
type_status <- legs_33 |>
  filter(!is.na(bill_source)) |>
  count(bill_source, bill_status) |>
  group_by(bill_source) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup()
```

```{r plot-bill-type, fig.width=7, fig.height=4.5}
outcome_palette <- c(
  "Enacted"   = "#2a9d8f",
  "Current"   = "#457b9d",
  "Defeated"  = "#e76f51",
  "Withdrawn" = "#f4a261",
  "Lapsed"    = "#bdbdbd"
)

ggplot(type_status,
       aes(x = pct, y = bill_source, fill = bill_status)) +
  geom_col(position = "stack", width = 0.5) +
  scale_fill_manual(values = outcome_palette, name = "Status") +
  scale_x_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(
    title    = "Bill Outcomes by Type",
    subtitle = "Government bills are far more likely to be enacted",
    x        = "Share of bills (%)",
    y        = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

---

## 5  Time from Introduction to Enactment

The API records `bill_year` (introduction) and `act_year` (enactment). Their
difference gives a coarse measure of legislative duration.

```{r enacted-duration}
enacted <- legs_33 |>
  filter(bill_status == "Enacted",
         !is.na(bill_year), !is.na(act_year)) |>
  mutate(
    year_introduced = as.integer(bill_year),
    year_enacted    = as.integer(act_year),
    years_to_enact  = year_enacted - year_introduced
  )

enacted |>
  summarise(
    n          = n(),
    min_yrs    = min(years_to_enact),
    median_yrs = median(years_to_enact),
    max_yrs    = max(years_to_enact)
  )
#> # A tibble: 1 × 4
#>       n min_yrs median_yrs max_yrs
#>   <int>   <int>      <dbl>   <int>
#> 1    94       0          1       4
```

```{r plot-duration, fig.width=6, fig.height=4}
ggplot(enacted, aes(x = years_to_enact, fill = bill_type)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(
    values = c("Government" = "#457b9d", "Private Members'" = "#e9c46a"),
    name   = "Bill type"
  ) +
  scale_x_continuous(breaks = 0:6) +
  labs(
    title    = "Years from Introduction to Enactment",
    subtitle = "Government bills typically pass within a year",
    x        = "Years",
    y        = "Number of bills"
  ) +
  theme_minimal(base_size = 13)
```

---

## 6  Most Prolific Sponsors

```{r top-sponsors}
top_sponsors <- legs_33 |>
  filter(!is.na(sponsors_name)) |>
  separate_longer_delim(sponsors_name,';')|>
  count(sponsors_name, bill_source, name = "n_bills") |>
  # Exclude generic government entries; keep individual names
  filter(!grepl("^Minister|^Government", sponsors_name)) |>
  slice_max(n_bills, n = 15) |>
  mutate(sponsors_name = reorder(sponsors_name, n_bills))
```

```{r plot-sponsors, fig.width=7, fig.height=6}
ggplot(top_sponsors,
       aes(x = n_bills, y = sponsors_name, fill = bill_source)) +
  geom_col(width = 0.65) +
  scale_fill_manual(
    values = c("Government" = "#457b9d", "Private Members'" = "#e9c46a"),
    name   = "Type"
  ) +
  labs(
    title = "Top 15 Bill Sponsors (33rd Dáil)",
    x     = "Bills introduced",
    y     = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

---

## 7  Tracking a Specific Bill

### 7.1  Look up by bill number and year

```{r single-bill}
# The Health (Miscellaneous Provisions) Bill 2022
hmp <- get_legislation(bill_no = 65, bill_year = 2022)

hmp |> select(title, bill_type, bill_status, sponsors_name,
              date_introduced, act_no, act_year)
#> # A tibble: 1 × 7
#>   title                           bill_type  bill_status sponsor_name
#>   <chr>                           <chr>      <chr>       <chr>
#> 1 Health (Miscellaneous Provis… Government  Enacted     Minister for Health
#>   date_introduced act_no act_year
#>   <chr>           <chr>  <chr>
#> 1 2022-10-18      6      2023
```

### 7.2  Find debates on the bill

Once you have the bill title, search for debates mentioning it. The
`get_debates()` endpoint can be filtered by date; combining that with
`get_debate_text()` lets you read the actual floor speeches.

```{r bill-debates}
# Debates from when the bill was active
bill_debates <- get_debates(
  chamber    = "dail",
  date_start = "2022-10-01",
  date_end   = "2023-06-30"
)

# The division subject field often contains the bill short title
bill_debates |>
  filter(grepl("Health.*Miscellaneous", section_title, ignore.case = TRUE)) |>
  select(date, section_title, formats_xml) |>
  head(5)
```

### 7.3  Read the transcript of the Second Stage debate

```{r bill-transcript}
# Use the formats_xml URI from the debates tibble
second_stage_uri <- bill_debates |>
  pull(formats_xml) |>
  head(1)

if (length(second_stage_uri) == 1 && !is.na(second_stage_uri)) {
  transcript <- get_debate_text(second_stage_uri)

  # Word count by speaker
  transcript |>
    mutate(word_count = lengths(strsplit(text, "\\s+"))) |>
    group_by(speaker_name) |>
    summarise(words = sum(word_count), speeches = n_distinct(speech_no)) |>
    arrange(desc(words)) |>
    head(10)
}
```

---

## 8  Linking Bills to Division Votes

Divisions (votes) are identified by their subject line, which usually contains
the bill short title and stage.

```{r find-divisions}
# All Dáil divisions in the period the bill was before the house
divs <- get_divisions(
  chamber    = "dail",
  date_start = "2022-10-01",
  date_end   = "2023-06-30"
)

# Filter to divisions that mention the bill
bill_votes <- divs |>
  filter(grepl("health", title, ignore.case = TRUE))

bill_votes |>
  select(date, vote_id, title, count_ta, count_nil, count_staon)
#> # A tibble: 4 × 6
#>   date       division_no subject                    count_ta count_nil count_staon
#>   <chr>            <chr> <chr>                         <chr>     <chr>       <chr>
#> 1 2023-04-18          43 Health (Misc Provs) Bill …       77        56           0
#> 2 2023-04-18          44 Health (Misc Provs) Bill …       79        55           0
#> ...
```

### 8.1  Vote breakdown for the final stage

```{r vote-breakdown}
# Take the last division (typically the 5th Stage / final vote)
final_div_id <- bill_votes |>
  slice_max(as.integer(gsub('vote_','',vote_id)), n = 1) |>
  pull(division_id)

votes <- get_division_votes(final_div_id)


votes <- bill_votes |>
  filter(division_id == final_div_id) |>
  select(division_id, date, member_id_ta, members_id_nil, members_id_staon) |>
  # 1. Turn the three columns into two: 'vote_type' and 'member_ids'
  tidyr::pivot_longer(
    cols = starts_with("member"), 
    names_to = "vote_type", 
    values_to = "member_id"
  ) |>
  # 2. Now separate. Since it's one column, different counts don't matter.
  tidyr::separate_longer_delim(member_id, delim = ";") |>
  # 3. Clean up the vote_type names (optional)
  mutate(vote_type = sub("members?_id_", "", vote_type)) |>
  # 4. Remove any empty rows (if a vote type had no members)
  filter(!is.na(member_id) & member_id != "")

print(votes)
votes |>
  count(vote_type, name = "n_members") |>
  mutate(vote = recode(vote_type, ta = "Tá (Yes)", nil = "Níl (No)",
                       staon = "Staon (Abstain)"))
#> # A tibble: 2 × 2
#>   vote          n_members
#>   <chr>             <int>
#> 1 Tá (Yes)             77
#> 2 Níl (No)             56
```

```{r plot-vote, fig.width=5, fig.height=3.5}
votes |>
  count(vote_type) |>
  mutate(
    vote  = recode(vote_type, ta = "Tá (Yes)", nil = "Níl (No)",
                   staon = "Staon (Abstain)"),
    vote  = factor(vote_type, levels = c("Tá (Yes)", "Níl (No)", "Staon (Abstain)"))
  ) |>
  ggplot(aes(x = vote, y = n, fill = vote)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.4, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c(
    "Tá (Yes)"       = "#2a9d8f",
    "Níl (No)"       = "#e76f51",
    "Staon (Abstain)"= "#bdbdbd"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Health (Misc. Provisions) Bill 2022 — Final Division",
    subtitle = paste0("Division no. "),
    x        = NULL,
    y        = "Members"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank())
```

---

## 9  Enacted Bills by Year

A simple time-series showing legislative output across years.

```{r enacted-all}
# Fetch all enacted bills (this is cached after the first call)
enacted_all <- get_legislation(bill_status = "Enacted", all_pages = TRUE)

enacted_by_year <- enacted_all |>
  filter(!is.na(act_year)) |>
  count(act_year, bill_type, name = "n_acts") |>
  mutate(act_year = as.integer(act_year)) |>
  filter(act_year >= 2010)
```

```{r plot-enacted-year, fig.width=8, fig.height=4.5}
ggplot(enacted_by_year,
       aes(x = act_year, y = n_acts, fill = bill_type)) +
  geom_col(position = "stack") +
  scale_fill_manual(
    values = c("Government" = "#457b9d", "Private Members'" = "#e9c46a"),
    name   = "Bill type"
  ) +
  scale_x_continuous(breaks = 2010:2025) +
  labs(
    title = "Acts of the Oireachtas by Year (2010–2024)",
    x     = NULL,
    y     = "Acts enacted"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## 10  A Complete Bill Audit: From Introduction to Transcript

Combining the four main functions into one analytical pipeline:

```{r full-audit}
## Step 1 – identify a bill
target <- get_legislation(bill_no = 105, bill_year = 2021) # e.g. Electoral Reform Bill

cat("Title:", target$title, "\n")
cat("Status:", target$bill_status, "\n")
cat("Introduced:", target$date_introduced, "\n")
cat("Enacted as:", paste0("No. ", target$act_no, " of ", target$act_year), "\n")

## Step 2 – find related debates
related <- get_debates(
  chamber    = "dail",
  date_start = target$date_introduced,
  date_end   = paste0(target$act_year, "-12-31")
)

bill_kw <- gsub("\\(.*\\)", "", target$title) |> trimws()   # strip parenthetical
bill_debates <- related |>
  filter(grepl(bill_kw, section_title, ignore.case = TRUE))

cat("Debate sections found:", nrow(bill_debates), "\n")

## Step 3 – fetch full transcript of the Second Stage debate
if (nrow(bill_debates) > 0) {
  transcript <- get_debate_text(bill_debates$formats_xml[1])

  word_share <- transcript |>
    mutate(words = lengths(strsplit(text, "\\s+"))) |>
    group_by(speaker_name) |>
    summarise(words = sum(words)) |>
    mutate(pct = round(words / sum(words) * 100, 1)) |>
    arrange(desc(words))

  print(word_share)
}

## Step 4 – find the final division vote
bill_divs <- get_divisions(
  chamber    = "dail",
  date_start = target$date_introduced,
  date_end   = paste0(target$act_year, "-12-31")
) |>
  filter(grepl(bill_kw, subject, ignore.case = TRUE))

if (nrow(bill_divs) > 0) {
  final_vote <- get_division_votes(tail(bill_divs$division_id, 1))
  cat("\nFinal division result:\n")
  print(count(final_vote, vote))
}
```

---

## Further Reading

- [Oireachtas Open Data API](https://api.oireachtas.ie/) — full endpoint
  reference and field descriptions.
- [Akoma Ntoso](http://www.akomantoso.org/) — the XML standard used for all
  debate transcripts, parseable with `get_debate_text()`.
- Other vignettes in this package:
  - *Sentiment analysis of debate transcripts*
  - *Topic modelling of parliamentary questions*
  - *Voting cohesion clustering*
  - *Member responsiveness around news events*
