---
title: "Member Profiling: Political Positioning via Question Analytics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Member Profiling: Political Positioning via Question Analytics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE,
  fig.width  = 9,
  fig.height = 7
)
```

> **Note:** All code chunks require a live internet connection. Fetching a
> year of written questions typically takes 3–5 seconds (results are cached).
> The PCA / UMAP steps run locally in milliseconds. UMAP requires the
> **uwot** package (`install.packages("uwot")`); all other packages are in the
> standard `oiReachtas` Suggests list.

---

## Moneyball for the Dáil

In basketball analytics a player's shot chart tells you *where* they score
from; a similarity algorithm then finds their nearest comps (e.g. "plays like
a young Ray Allen"). In this vignette we apply exactly the same logic to Dáil
members:

| Sports concept | Political equivalent |
|---|---|
| Shot-chart / touch heat-map | **TF-IDF question fingerprint** |
| Positional clustering | **Policy-domain radar** |
| Dimensionality reduction (PCA / UMAP) | **Political map** (left–right, activist–loyalist) |
| Player archetypes (stretch big, point forward …) | **Member archetypes** (k-means clusters) |
| "Comps" (most similar players) | **Member comps** (cosine similarity) |
| Stats radar / scouting card | **Member scouting card** |

The raw material is every **written question** a member has tabled — the
show-as title is a short, clean summary of the topic they chose to ask about.
This corpus is a revealed-preference record: members decide what to ask about,
so it reflects genuine priority rather than party press-release language.

---

## 1  Setup

```{r libraries}
library(oiReachtas)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(ggrepel)
library(scales)
library(tidytext)   # TF-IDF, tokenisation

# UMAP — install if needed: install.packages("uwot")
library(uwot)
```

---

## 2  Fetch the Question Corpus

We pull written questions for the most recent completed Dáil session and
attach party and constituency metadata.

```{r fetch-questions}
# ── member roster (names, party, constituency) ──────────────────────────────
members <- get_members(house = "dail", limit = 200) |>
  select(member_id  = member_pId,
         member_name,
         party       = party_code,
         constituency)

# ── 12 months of written questions ──────────────────────────────────────────
qs_raw <- get_questions(
  date_start  = "2024-01-01",
  date_end    = "2024-12-31",
  question_type = "written",
  limit       = 5000
)

# Keep only the fields we need and join member metadata
qs <- qs_raw |>
  select(member_pId, show_as, dept_name) |>
  left_join(members, by = c("member_pId" = "member_id")) |>
  filter(!is.na(member_name), !is.na(show_as), nchar(show_as) > 5)

glimpse(qs)
```

---

## 3  TF-IDF Question Fingerprints — the "Touch Heat-Map"

For each member we tokenise every question title, remove stop-words
(English + Oireachtas-specific), and compute **TF-IDF** scores.
The result is a member × vocabulary matrix: high TF-IDF means a word is
distinctive to *that* member compared with the rest of the Dáil.

```{r tfidf}
# Oireachtas-specific stop words (procedural boilerplate)
oireachtas_stops <- tibble(word = c(
  "minister", "government", "plans", "measures", "steps",
  "action", "actions", "information", "details", "aware",
  "issues", "concerned", "basis", "provision", "provisions",
  "matter", "matters", "support", "supports", "service", "services",
  "scheme", "schemes", "regard", "relevant", "area", "areas",
  "funding", "ensure", "provide", "policy", "policies", "report",
  "request", "asked", "stated", "confirm", "outlined"
))

# Tokenise + TF-IDF ──────────────────────────────────────────────────────────
tidy_qs <- qs |>
  unnest_tokens(word, show_as) |>
  anti_join(stop_words,       by = "word") |>
  anti_join(oireachtas_stops, by = "word") |>
  filter(nchar(word) > 2, !grepl("^[0-9]+$", word))

member_tfidf <- tidy_qs |>
  count(member_name, word, name = "n") |>
  bind_tf_idf(term = word, document = member_name, n = n)

# Top-5 signature terms per member (their "shot chart")
member_tfidf |>
  group_by(member_name) |>
  slice_max(tf_idf, n = 5, with_ties = FALSE) |>
  ungroup() |>
  arrange(member_name) |>
  print(n = 30)
```

### 3.1  Party-level vocabulary

Before individual profiling, see which terms most distinguish each party.

```{r party-vocab}
party_tfidf <- tidy_qs |>
  left_join(select(qs, member_name, party), by = "member_name") |>
  count(party, word, name = "n") |>
  bind_tf_idf(term = word, document = party, n = n)

party_tfidf |>
  group_by(party) |>
  slice_max(tf_idf, n = 6, with_ties = FALSE) |>
  ungroup() |>
  mutate(word = reorder_within(word, tf_idf, party)) |>
  ggplot(aes(tf_idf, word, fill = party)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~party, scales = "free_y", ncol = 3) +
  scale_y_reordered() +
  scale_fill_manual(values = c(
    "FF" = "#66BB6A", "FG" = "#42A5F5", "SF" = "#4CAF50",
    "Lab" = "#EF5350", "GP" = "#26A69A", "SD" = "#AB47BC",
    "PBP" = "#FF7043", "Ind" = "#78909C"
  ), na.value = "#BDBDBD") +
  labs(
    title    = "Party political vocabulary",
    subtitle = "Top TF-IDF terms per party — what distinguishes each party's questions",
    x        = "TF-IDF score", y = NULL,
    caption  = "Source: Oireachtas Open Data API"
  ) +
  theme_minimal(base_size = 11)
```

---

## 4  Policy-Domain Radar — the "Scouting Card"

We map vocabulary onto ten policy domains and compute each member's
question-share per domain.  The resulting radar is like an NBA stats radar:
a snapshot of where a member invests their parliamentary attention.

```{r domain-keywords}
domains <- list(
  "Housing & Planning"    = c("housing", "rent", "home", "tenant", "landlord",
                               "accommodation", "planning", "homeless", "eviction",
                               "hap", "hal", "rents", "residential", "development"),
  "Health & HSE"          = c("health", "hospital", "hse", "waiting", "gp",
                               "mental", "nursing", "cancer", "ambulance", "pharmacy",
                               "disability", "maternity", "beds", "staffing"),
  "Education"             = c("school", "teacher", "education", "college",
                               "university", "student", "special", "sen", "curriculum",
                               "principal", "snc", "sna", "class", "pupil"),
  "Transport"             = c("road", "bus", "train", "rail", "cycle", "traffic",
                               "transport", "motorway", "greenway", "brt", "luas",
                               "dart", "airport", "port"),
  "Environment & Climate" = c("climate", "carbon", "emission", "renewable",
                               "energy", "wind", "solar", "waste", "water", "flood",
                               "biodiversity", "nature", "offshore", "fossil"),
  "Social Welfare"        = c("welfare", "pension", "disability", "carer",
                               "benefit", "payment", "allowance", "dsp", "social",
                               "lone", "jobseeker", "community", "poverty"),
  "Justice & Crime"       = c("garda", "crime", "court", "prison", "legal",
                               "justice", "solicitor", "judge", "offence", "victim",
                               "drugs", "trafficking", "violence", "fraud"),
  "Agriculture & Rural"   = c("farmer", "farm", "agri", "dairy", "beef", "rural",
                               "fishing", "forestry", "land", "mart", "tillage",
                               "dafm", "animal", "felling"),
  "Enterprise & Economy"  = c("enterprise", "business", "sme", "employment",
                               "job", "tax", "vat", "trade", "investment", "startup",
                               "export", "ibec", "isme", "commercial"),
  "Foreign & Defence"     = c("foreign", "ukraine", "conflict", "israel",
                               "palestine", "nato", "defence", "international",
                               "humanitarian", "migration", "asylum", "refugee",
                               "embassy", "united")
)

# Build domain lookup table
domain_lookup <- imap_dfr(domains, ~ tibble(word = .x, domain = .y))

# Score each member per domain
member_domains <- tidy_qs |>
  inner_join(domain_lookup, by = "word") |>
  count(member_name, domain, name = "domain_n") |>
  group_by(member_name) |>
  mutate(domain_pct = domain_n / sum(domain_n) * 100) |>
  ungroup()

# Summary: which domains dominate for each party?
member_domains |>
  left_join(select(qs, member_name, party) |> distinct(), by = "member_name") |>
  group_by(party, domain) |>
  summarise(mean_pct = mean(domain_pct), .groups = "drop") |>
  slice_max(mean_pct, n = 3, by = party) |>
  arrange(party, desc(mean_pct))
```

### 4.1  Radar chart for individual members

```{r radar-fn}
# Helper: draw a ggplot2 polar radar for a given member
plot_member_radar <- function(member, colour = "#42A5F5") {
  # all 10 domains, fill 0 if member never asked about one
  d <- tibble(domain = names(domains)) |>
    left_join(
      filter(member_domains, member_name == member),
      by = "domain"
    ) |>
    mutate(domain_pct = replace_na(domain_pct, 0))

  # Shorten labels for display
  d$label <- stringr::str_wrap(d$domain, width = 12)

  ggplot(d, aes(x = domain, y = domain_pct)) +
    geom_col(fill = colour, alpha = 0.75, width = 0.85) +
    coord_polar(theta = "x", start = -pi / length(domains)) +
    scale_y_continuous(labels = function(x) paste0(x, "%")) +
    labs(
      title    = paste("Policy radar —", member),
      subtitle = "% of domain-tagged questions per policy area",
      x = NULL, y = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x  = element_text(size = 8, colour = "grey30"),
      axis.text.y  = element_text(size = 7, colour = "grey50"),
      panel.grid   = element_line(colour = "grey88"),
      plot.title   = element_text(face = "bold")
    )
}

# ── Example: compare two members side by side ───────────────────────────────
# Replace with any names returned by get_members()
p1 <- plot_member_radar("Mary Lou McDonald", colour = "#4CAF50")
p2 <- plot_member_radar("Micheál Martin",    colour = "#66BB6A")

# Combine with patchwork if available, otherwise print individually
if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p1, p2)
} else {
  print(p1)
  print(p2)
}
```

---

## 5  Political Map — PCA Projection

We cast the TF-IDF scores into a **member × term** matrix and apply PCA.
The first two principal components typically correspond to:

* **PC1** — a government/opposition axis (usually explains 15–25 % of variance)
* **PC2** — a urban-activist vs. rural-traditionalist axis

This is the political equivalent of projecting players into a 2D "skill
space" using advanced metrics.

```{r pca-matrix}
# ── Wide matrix: rows = members, columns = terms ────────────────────────────
# We use TF-IDF score; keep only terms appearing for ≥ 3 members (sparsity)
common_terms <- member_tfidf |>
  count(word) |>
  filter(n >= 3) |>
  pull(word)

member_mat <- member_tfidf |>
  filter(word %in% common_terms) |>
  select(member_name, word, tf_idf) |>
  pivot_wider(names_from = word, values_from = tf_idf, values_fill = 0) |>
  column_to_rownames("member_name")

cat("Matrix dimensions:", nrow(member_mat), "members ×",
    ncol(member_mat), "terms\n")
```

```{r pca}
pca_res <- prcomp(member_mat, center = TRUE, scale. = TRUE)

# Scree plot — how many components are meaningful?
tibble(
  pc  = seq_along(pca_res$sdev),
  var = pca_res$sdev^2 / sum(pca_res$sdev^2) * 100
) |>
  filter(pc <= 15) |>
  ggplot(aes(pc, var)) +
  geom_col(fill = "#5C6BC0") +
  geom_line(colour = "grey40") +
  geom_point(colour = "grey40") +
  labs(
    title = "PCA scree plot",
    x     = "Principal component",
    y     = "Variance explained (%)"
  ) +
  theme_minimal()
```

```{r pca-map}
# PC scores + party labels
pca_scores <- as_tibble(pca_res$x[, 1:4], rownames = "member_name") |>
  left_join(select(members, member_name, party, constituency), by = "member_name")

# Party palette
party_colours <- c(
  "FF"  = "#66BB6A", "FG" = "#42A5F5", "SF" = "#4CAF50",
  "Lab" = "#EF5350", "GP" = "#26A69A", "SD" = "#AB47BC",
  "PBP" = "#FF7043", "Ind" = "#78909C"
)

ggplot(pca_scores, aes(PC1, PC2, colour = party, label = member_name)) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_text_repel(size = 2.5, max.overlaps = 18, segment.colour = "grey70") +
  scale_colour_manual(values = party_colours, na.value = "#BDBDBD") +
  labs(
    title    = "Dáil members — political map (question-based PCA)",
    subtitle = "PC1 ≈ government–opposition axis · PC2 ≈ activist–rural axis",
    x        = "PC1", y = "PC2", colour = "Party",
    caption  = "Source: Oireachtas Open Data API — written questions 2024"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

### 5.1  What drives each axis?

Inspect the **loadings** to understand what PC1 and PC2 actually measure.

```{r pca-loadings}
loadings <- as_tibble(pca_res$rotation[, 1:2], rownames = "word")

# Top 15 terms pulling in each direction
bind_rows(
  slice_min(loadings, PC1, n = 15) |> mutate(direction = "← PC1 (govt end)"),
  slice_max(loadings, PC1, n = 15) |> mutate(direction = "→ PC1 (oppo end)")
) |>
  mutate(word = reorder(word, PC1)) |>
  ggplot(aes(PC1, word, fill = direction)) +
  geom_col() +
  scale_fill_manual(values = c("← PC1 (govt end)" = "#42A5F5",
                               "→ PC1 (oppo end)" = "#EF5350")) +
  labs(title = "PC1 loadings — what defines the main axis?",
       x = "Loading", y = NULL, fill = NULL) +
  theme_minimal(base_size = 11)
```

---

## 6  UMAP — Finer Cluster Structure

PCA is linear; UMAP preserves local neighbourhood structure and often reveals
sub-clusters that PCA collapses. Think of it as going from a basic positional
map to a detailed player-tracking heat-map.

```{r umap}
# Requires uwot — install.packages("uwot")
set.seed(42)

umap_coords <- uwot::umap(
  member_mat,
  n_neighbors = 12,     # local neighbourhood size
  min_dist    = 0.15,   # how tightly to pack clusters
  metric      = "cosine"
)

umap_df <- tibble(
  member_name = rownames(member_mat),
  UMAP1       = umap_coords[, 1],
  UMAP2       = umap_coords[, 2]
) |>
  left_join(select(members, member_name, party, constituency), by = "member_name")

ggplot(umap_df, aes(UMAP1, UMAP2, colour = party, label = member_name)) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_text_repel(size = 2.5, max.overlaps = 20, segment.colour = "grey70") +
  scale_colour_manual(values = party_colours, na.value = "#BDBDBD") +
  labs(
    title    = "Dáil members — UMAP projection (cosine distance on question TF-IDF)",
    subtitle = "Proximity = similar question portfolio; colour = party",
    x = "UMAP 1", y = "UMAP 2", colour = "Party",
    caption  = "Source: Oireachtas Open Data API — written questions 2024"
  ) +
  theme_minimal(base_size = 11)
```

---

## 7  Member Archetypes — K-Means Clustering

K-means on the first 10 PC scores partitions members into archetypes.
We choose *k* = 6 (tune with the elbow method below) and then characterise
each cluster by its top discriminating terms — naming the "player types".

```{r elbow}
# Elbow method — total within-cluster SS vs. k
pc_scores_mat <- pca_res$x[, 1:10]

wss <- map_dbl(1:10, ~ {
  kmeans(pc_scores_mat, centers = .x, nstart = 20)$tot.withinss
})

tibble(k = 1:10, wss = wss) |>
  ggplot(aes(k, wss)) +
  geom_line(colour = "#5C6BC0") +
  geom_point(colour = "#5C6BC0", size = 3) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Elbow method — choosing k for member archetypes",
    x     = "Number of clusters (k)",
    y     = "Total within-cluster SS"
  ) +
  theme_minimal()
```

```{r kmeans}
set.seed(42)
k_best   <- 6   # adjust after inspecting the elbow plot
km_res   <- kmeans(pc_scores_mat, centers = k_best, nstart = 25, iter.max = 50)

cluster_df <- tibble(
  member_name = rownames(member_mat),
  archetype   = factor(km_res$cluster)
) |>
  left_join(select(members, member_name, party), by = "member_name")

# Party composition of each archetype
cluster_df |>
  count(archetype, party) |>
  group_by(archetype) |>
  mutate(pct = n / sum(n) * 100) |>
  slice_max(pct, n = 3) |>
  arrange(archetype, desc(pct))
```

### 7.1  Name the archetypes

Inspect each cluster's top TF-IDF terms to assign a descriptive label.

```{r archetype-names}
# Average TF-IDF per cluster per word
archetype_terms <- member_tfidf |>
  filter(word %in% common_terms) |>
  left_join(select(cluster_df, member_name, archetype), by = "member_name") |>
  group_by(archetype, word) |>
  summarise(mean_tfidf = mean(tf_idf), .groups = "drop")

# Top 8 words per archetype
archetype_terms |>
  group_by(archetype) |>
  slice_max(mean_tfidf, n = 8, with_ties = FALSE) |>
  ungroup() |>
  mutate(word = reorder_within(word, mean_tfidf, archetype)) |>
  ggplot(aes(mean_tfidf, word, fill = archetype)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~archetype, scales = "free_y", ncol = 3,
             labeller = as_labeller(c(
               "1" = "Archetype 1", "2" = "Archetype 2",
               "3" = "Archetype 3", "4" = "Archetype 4",
               "5" = "Archetype 5", "6" = "Archetype 6"
             ))) +
  scale_y_reordered() +
  labs(
    title    = "Member archetypes — defining vocabulary",
    subtitle = "Mean TF-IDF per cluster; label each archetype from these terms",
    x        = "Mean TF-IDF", y = NULL
  ) +
  theme_minimal(base_size = 10)
```

```{r archetype-map}
# Overlay archetypes on the UMAP map
umap_df |>
  left_join(select(cluster_df, member_name, archetype), by = "member_name") |>
  ggplot(aes(UMAP1, UMAP2, colour = archetype, label = member_name)) +
  geom_point(size = 3, alpha = 0.85) +
  geom_text_repel(size = 2.4, max.overlaps = 16, segment.colour = "grey70") +
  scale_colour_brewer(palette = "Set2") +
  labs(
    title    = "Member archetypes on the UMAP map",
    x = "UMAP 1", y = "UMAP 2", colour = "Archetype"
  ) +
  theme_minimal(base_size = 11)
```

---

## 8  Player Comps — Cosine Similarity Search

The cosine similarity between two TF-IDF vectors measures how alike two
members' question portfolios are, regardless of overall volume.
A score of 1 = identical distribution; 0 = completely different focus.

```{r cosine}
# Row-normalise the TF-IDF matrix → unit vectors → dot product = cosine sim
mat_norm  <- member_mat / sqrt(rowSums(member_mat^2) + 1e-10)
cosine_sim <- as.matrix(mat_norm) %*% t(as.matrix(mat_norm))

# Helper: return top-n comps for a given member
find_comps <- function(member, n = 5) {
  if (!member %in% rownames(cosine_sim)) {
    stop("Member not found. Check spelling against get_members() output.")
  }
  sims <- sort(cosine_sim[member, ], decreasing = TRUE)
  sims <- sims[names(sims) != member]   # remove self
  tibble(
    comp       = names(head(sims, n)),
    similarity = round(head(sims, n), 3)
  ) |>
    left_join(select(members, member_name, party, constituency),
              by = c("comp" = "member_name"))
}

# ── Example: find comps for two members ─────────────────────────────────────
find_comps("Mary Lou McDonald")
find_comps("Micheál Martin")
```

### 8.1  Full similarity network (top comps per member)

```{r sim-network}
# Build edges: each member → their top-3 comps
edges <- rownames(cosine_sim) |>
  map_dfr(\(m) {
    sims <- sort(cosine_sim[m, ], decreasing = TRUE)
    sims <- sims[names(sims) != m]
    tibble(from = m, to = names(head(sims, 3)),
           weight = head(sims, 3))
  })

# Visualise on the UMAP layout — draw lines between comps
umap_coords_named <- umap_df |>
  select(member_name, UMAP1, UMAP2)

edges_with_coords <- edges |>
  left_join(umap_coords_named, by = c("from" = "member_name")) |>
  rename(x = UMAP1, y = UMAP2) |>
  left_join(umap_coords_named, by = c("to" = "member_name"),
            suffix = c("", "_end")) |>
  rename(xend = UMAP1_end, yend = UMAP2_end)

ggplot() +
  geom_segment(data = edges_with_coords,
               aes(x = x, y = y, xend = xend, yend = yend, alpha = weight),
               colour = "#5C6BC0", linewidth = 0.4) +
  geom_point(data = umap_df,
             aes(UMAP1, UMAP2, colour = party), size = 2.5) +
  scale_colour_manual(values = party_colours, na.value = "#BDBDBD") +
  scale_alpha_continuous(range = c(0.1, 0.6), guide = "none") +
  labs(
    title    = "Member similarity network (top-3 comps per member)",
    subtitle = "Lines connect members with most similar question portfolios",
    x = "UMAP 1", y = "UMAP 2", colour = "Party"
  ) +
  theme_minimal(base_size = 11)
```

---

## 9  Left–Right Positioning

Two complementary approaches; combining them gives a more robust score.

### 9.1  Text-based left–right score

We score each question word as *left-leaning* or *right-leaning* using a
small seed dictionary and compute a net score per member.

```{r lr-text}
left_words  <- c("equality", "workers", "union", "welfare", "public",
                 "affordable", "redistribute", "poverty", "rights",
                 "universal", "social", "care", "community", "homeless")
right_words <- c("enterprise", "business", "market", "private", "tax",
                 "regulation", "deregulation", "competition", "investment",
                 "commercial", "fiscal", "property", "landlord", "growth")

lr_scores_text <- tidy_qs |>
  mutate(signal = case_when(
    word %in% left_words  ~ -1,
    word %in% right_words ~  1,
    TRUE                  ~  0
  )) |>
  filter(signal != 0) |>
  group_by(member_name) |>
  summarise(lr_text = mean(signal), .groups = "drop")

# Merge with party labels and sort
lr_scores_text |>
  left_join(select(members, member_name, party), by = "member_name") |>
  arrange(lr_text) |>
  mutate(member_name = factor(member_name, levels = member_name)) |>
  filter(!is.na(lr_text)) |>
  ggplot(aes(lr_text, member_name, colour = party)) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = party_colours, na.value = "#BDBDBD") +
  labs(
    title    = "Text-based left–right score",
    subtitle = "Negative = left-leaning vocabulary · Positive = right-leaning vocabulary",
    x        = "Mean left–right signal", y = NULL, colour = "Party"
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_text(size = 7))
```

### 9.2  Vote-based left–right score (PC1 from divisions)

Cross-reference with the vote-based axis computed in the *Voting Cohesion*
vignette.  Load a saved RDS or recompute here:

```{r lr-votes}
# Assumes you have run the voting-cohesion vignette and saved:
#   saveRDS(vote_pca_scores, "data/vote_pca_scores.rds")
# Adjust path as needed.

vote_pc1 <- tryCatch(
  readRDS("data/vote_pca_scores.rds") |>
    select(member_name, lr_vote = PC1),
  error = function(e) {
    message("Vote PCA scores not found. Run the voting-cohesion vignette first.")
    NULL
  }
)

if (!is.null(vote_pc1)) {
  # Compare text-based vs vote-based positioning
  combined_lr <- lr_scores_text |>
    inner_join(vote_pc1, by = "member_name") |>
    left_join(select(members, member_name, party), by = "member_name")

  ggplot(combined_lr, aes(lr_vote, lr_text, colour = party,
                          label = member_name)) +
    geom_point(size = 2.5, alpha = 0.85) +
    geom_text_repel(size = 2.4, max.overlaps = 15) +
    geom_smooth(method = "lm", se = TRUE, colour = "grey40",
                linetype = "dashed") +
    scale_colour_manual(values = party_colours, na.value = "#BDBDBD") +
    labs(
      title    = "Text-based vs vote-based left–right positioning",
      subtitle = "Strong correlation validates both measures",
      x        = "Vote PC1 (govt → opposition)",
      y        = "Text LR score (left → right)",
      colour   = "Party"
    ) +
    theme_minimal(base_size = 11)
}
```

---

## 10  The Member Scouting Card

Bringing it all together: a single function that prints a full member profile
— archetype, top terms, domain radar, political position, and five comps.

```{r scouting-card}
member_scouting_card <- function(member, archetype_labels = NULL) {
  # ── sanity check ──────────────────────────────────────────────────────────
  if (!member %in% rownames(member_mat)) {
    stop("Member '", member, "' not found in the question corpus.")
  }

  cat("═══════════════════════════════════════════════════════════\n")
  cat("  MEMBER SCOUTING CARD\n")
  cat("  ", member, "\n")
  cat("═══════════════════════════════════════════════════════════\n\n")

  # Metadata
  meta <- members |> filter(member_name == member)
  cat("  Party:          ", meta$party %||% "—", "\n")
  cat("  Constituency:   ", meta$constituency %||% "—", "\n")

  # Archetype
  arch <- cluster_df |> filter(member_name == member) |> pull(archetype)
  arch_label <- if (!is.null(archetype_labels)) archetype_labels[arch] else
                paste("Archetype", arch)
  cat("  Archetype:      ", arch_label, "\n\n")

  # Top 8 TF-IDF terms
  top_terms <- member_tfidf |>
    filter(member_name == member) |>
    slice_max(tf_idf, n = 8, with_ties = FALSE) |>
    pull(word)
  cat("  Signature terms:", paste(top_terms, collapse = ", "), "\n\n")

  # Political position
  lr <- lr_scores_text |> filter(member_name == member) |> pull(lr_text)
  pos_label <- case_when(
    lr < -0.3 ~ "Left",
    lr < -0.1 ~ "Centre-left",
    lr <  0.1 ~ "Centre",
    lr <  0.3 ~ "Centre-right",
    TRUE      ~ "Right"
  )
  cat("  Political position (text-based):", pos_label,
      sprintf("(score: %.3f)\n\n", lr))

  # Top-3 policy domains
  top_domains <- member_domains |>
    filter(member_name == member) |>
    slice_max(domain_pct, n = 3) |>
    mutate(label = sprintf("%s (%.0f%%)", domain, domain_pct)) |>
    pull(label)
  cat("  Top policy domains:\n")
  for (d in top_domains) cat("    ·", d, "\n")
  cat("\n")

  # Player comps
  comps <- find_comps(member, n = 5)
  cat("  Most similar members (comps):\n")
  for (i in seq_len(nrow(comps))) {
    cat(sprintf("    %d. %-30s [%s] sim = %.3f\n",
                i, comps$comp[i],
                comps$party[i] %||% "—",
                comps$similarity[i]))
  }
  cat("\n")

  # Print radar chart
  print(plot_member_radar(member))

  invisible(list(
    meta         = meta,
    archetype    = arch,
    top_terms    = top_terms,
    lr_score     = lr,
    top_domains  = top_domains,
    comps        = comps
  ))
}

# Null-coalescing operator (base R doesn't have one)
`%||%` <- function(a, b) if (!is.null(a) && length(a) && !is.na(a[1])) a else b

# ── Run examples ─────────────────────────────────────────────────────────────
member_scouting_card("Mary Lou McDonald")
member_scouting_card("Micheál Martin")
```

---

## 11  Extending to Debate Transcripts

Written questions capture *what* members ask about; **debate contributions**
reveal *how* they argue.  `get_debate_text()` returns the raw speech for any
debate record, giving a much richer text corpus for the same pipeline.

```{r debates-extension}
# Fetch a sample of recent debates
recent_debates <- get_debates(
  date_start = "2024-10-01",
  date_end   = "2024-12-31",
  limit      = 200
)

# Pull speech text per member (this takes a few minutes for a large corpus)
debate_speeches <- recent_debates |>
  slice_head(n = 50) |>        # start small; expand as needed
  pull(debate_id) |>
  map_dfr(\(id) {
    tryCatch(
      get_debate_text(id) |> mutate(debate_id = id),
      error = function(e) NULL
    )
  })

# debate_speeches now has member_name + text columns.
# Feed into the same TF-IDF pipeline above (replace `show_as` with `text`)
# to get a richer fingerprint that blends questions + speeches.
#
# Tip: weight question TF-IDF and speech TF-IDF separately, then combine:
#   combined_tfidf <- bind_rows(
#     question_tfidf |> mutate(source = "question", weight = 1.0),
#     speech_tfidf   |> mutate(source = "speech",   weight = 0.6)
#   ) |>
#   mutate(tf_idf_w = tf_idf * weight)
```

---

## Summary

| Step | Method | What it tells you |
|------|--------|--------------------|
| TF-IDF fingerprint | `bind_tf_idf()` | A member's signature vocabulary |
| Domain radar | keyword → domain lookup | Which policy areas they focus on |
| PCA political map | `prcomp()` | Left–right + activist–rural positioning |
| UMAP | `uwot::umap()` | Fine-grained neighbourhood clusters |
| Archetypes | k-means on PC scores | Canonical "player types" in the Dáil |
| Player comps | cosine similarity | Most similar members by question portfolio |
| Scouting card | combined output | Full member profile in one call |

The pipeline is entirely data-driven — the political labels emerge from the
text, not from assumptions.  Feed in a new Dáil term, a different house
(Seanad), or a different country's open parliamentary data and the analysis
re-runs unchanged.
