---
title: "Ministerial Responsiveness: Deflection, Verbosity, and Clarity in Oral Questions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ministerial Responsiveness: Deflection, Verbosity, and Clarity in Oral Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```

> **Note:** All code chunks require a live internet connection. Fetching
> three months of oral-question debate transcripts typically takes 20–40
> seconds (full-text XML is retrieved per debate). All intermediate objects
> are built in-session; re-run `fetch-debates` and `fetch-text` chunks to
> refresh the corpus.

---

## What Makes a Good Ministerial Answer?

When a TD asks an oral question in the Dáil, the minister's reply can be
evaluated on at least three measurable dimensions that together capture the
quality of democratic accountability:

| Dimension | What it measures | Good score |
|---|---|---|
| **Verbosity** | Word count of the answer | Moderate — too short is dismissive, too long buries the point |
| **Deflection** | Cosine similarity between question TF-IDF and answer TF-IDF | High — a responsive answer shares topic vocabulary with the question |
| **Clarity** | Flesch-Kincaid readability of the answer | High — accessible prose scores higher |

Together these three axes produce a *ministerial responsiveness scorecard*:
a minister who gives long, jargon-heavy answers that are topically unrelated
to the question asked is deflecting through verbosity and complexity.

```
  ┌──────────────────────────────────────────────────────────────────┐
  │  Ideal answer                                                    │
  │     • moderate length   (not one-line brush-off, not wall of text)│
  │     • high cosine sim   (vocabulary overlaps with the question)   │
  │     • high readability  (Flesch-Kincaid ≥ 40)                    │
  └──────────────────────────────────────────────────────────────────┘
```

---

## 1  Setup

```{r libraries}
library(oiReachtas)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(ggrepel)
library(scales)
library(tidytext)

# Party colour palette ── used throughout for consistent visual identity
party_colours <- c(
  FF  = "#66BB6A",
  FG  = "#42A5F5",
  SF  = "#4CAF50",
  Lab = "#EF5350",
  GP  = "#26A69A",
  SD  = "#AB47BC",
  PBP = "#FF7043",
  Ind = "#78909C"
)
```

---

## 2  Fetch Oral Question Debate Transcripts

We use a three-month window. `get_debates()` returns one row per debate
section; those with `debate_type == "oral_questions"` contain the Q&A
exchanges we want. The `formats_xml` column holds the URI we pass to
`get_debate_text()` to retrieve the full transcript.

```{r fetch-debates}
# ── 1.  Discover oral-question debates ──────────────────────────────────────
debates_raw <- get_debates(
  chamber    = "dail",
  date_start = "2024-01-01",
  date_end   = "2024-03-31",
  limit      = 1000,
  all_pages  = TRUE
)

# Keep only oral-question sections that have an XML transcript URI
oral_debates <- debates_raw |>
  filter(debate_type == "oral_questions",
         !is.na(formats_xml),
         formats_xml != "")

cat(sprintf("Found %d oral-question debate sections.\n", nrow(oral_debates)))
```

```{r fetch-text}
# ── 2.  Retrieve full transcript for each debate section ────────────────────
#
#  get_debate_text() is called with by_para = FALSE so we get one row per
#  speech rather than one row per paragraph — easier to pair Q with A.
#
#  We use purrr::map() over formats_xml and bind results together.

fetch_safe <- possibly(
  .f = function(uri) get_debate_text(uri, by_para = FALSE),
  otherwise = NULL
)

transcripts_raw <- debates_raw |>
  select(debate_id, date, formats_xml) |>
  mutate(
    transcript = map(formats_xml, fetch_safe)
  ) |>
  filter(!map_lgl(transcript, is.null)) |>
  unnest(transcript)

# Confirm columns present
glimpse(transcripts_raw)
# Expected: section_id, section_heading, speech_no, speech_type,
#           speaker_ref, speaker_uri, speaker_name, para_no, text
```

---

## 3  Pair Questions with Answers

Each oral question and its corresponding ministerial answer share the same
`section_id`. We split the transcript into question speeches and answer
speeches, then join them by `section_id` so every row contains both the
question text and the answer text for that exchange.

```{r pair-qa}
# ── Keep only question and answer speech types ───────────────────────────────
qa_speeches <- transcripts_raw |>
  filter(speech_type %in% c("question", "speech"),
         !is.na(text),
         nchar(trimws(text)) > 0)

# ── Split into questions and answers ─────────────────────────────────────────
questions_df <- qa_speeches |>
  filter(speech_type == "question") |>
  group_by(section_id) |>
  summarise(
    question_text  = paste(text, collapse = " "),
    questioner     = first(speaker_name),
    .groups        = "drop"
  )

answers_df <- qa_speeches |>
  filter(speech_type == "speech") |>
  group_by(section_id) |>
  summarise(
    answer_text    = paste(text, collapse = " "),
    minister_name  = first(speaker_name),
    minister_ref   = first(speaker_ref),
    .groups        = "drop"
  )

# ── Join on section_id; attach date and department metadata ──────────────────
qa_pairs <- questions_df |>
  inner_join(answers_df, by = "section_id") |>
  left_join(
    transcripts_raw |>
      distinct(section_id, date, section_heading, debate_id),
    by = "section_id"
  ) |>
  # Extract department from section_heading (typically "Minister for X")
  mutate(
    department = stringr::str_extract(
      section_heading,
      "(?<=Minister for )([^–—-]+)"
    ) |> trimws()
  )

cat(sprintf("Paired %d question–answer exchanges across %d debate sections.\n",
            nrow(qa_pairs),
            n_distinct(qa_pairs$debate_id)))
```

---

## 4  Answer Length: Verbosity by Department

Word count is the most direct measure of how much (or how little) a minister
actually said. A violin plot shows the full distribution within each
department, revealing whether verbosity is minister-specific or department-
wide.

```{r answer-length}
# ── Compute word counts ───────────────────────────────────────────────────────
qa_pairs <- qa_pairs |>
  mutate(
    answer_words   = stringr::str_count(answer_text,   "\\S+"),
    question_words = stringr::str_count(question_text, "\\S+")
  )

# ── Summarise at department level for ordering ────────────────────────────────
dept_order <- qa_pairs |>
  filter(!is.na(department)) |>
  group_by(department) |>
  summarise(median_words = median(answer_words, na.rm = TRUE),
            .groups = "drop") |>
  arrange(median_words) |>
  pull(department)

plot_verbosity <- qa_pairs |>
  filter(!is.na(department),
         answer_words > 0,
         answer_words < quantile(answer_words, 0.99, na.rm = TRUE)) |>
  mutate(department = factor(department, levels = dept_order))

ggplot(plot_verbosity, aes(x = department, y = answer_words)) +
  geom_violin(fill = "#42A5F5", alpha = 0.6, colour = "white", scale = "width") +
  geom_boxplot(width = 0.12, outlier.shape = NA,
               fill = "white", colour = "#1565C0", linewidth = 0.4) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Ministerial Answer Length by Department",
    subtitle = "Oral questions, Dáil Éireann · Jan–Mar 2024",
    x        = NULL,
    y        = "Words in ministerial answer",
    caption  = "Source: Oireachtas Open Data API via oiReachtas"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.title         = element_text(face = "bold")
  )
```

---

## 5  Deflection Score: Cosine Similarity of Question vs Answer

A minister who actually answers the question will use vocabulary related to
what was asked. We measure this with **cosine similarity** between the TF-IDF
vectors of the question text and the answer text within the same `section_id`.

A score near **1** means the answer is topically aligned with the question;
a score near **0** means the answer shares almost no vocabulary with what was
asked — a strong signal of deflection.

```{r cosine-sim-helper}
# ── Cosine similarity helper ──────────────────────────────────────────────────
#
#  Inputs:  two named numeric vectors (TF-IDF weights, names = terms)
#  Output:  scalar in [0, 1]

cosine_sim <- function(a, b) {
  # Keep only terms that appear in both vectors
  common <- intersect(names(a), names(b))
  if (length(common) == 0L) return(0)
  a_c  <- a[common]
  b_c  <- b[common]
  dot  <- sum(a_c * b_c)
  norm <- sqrt(sum(a^2)) * sqrt(sum(b^2))
  if (norm == 0) return(0)
  dot / norm
}
```

```{r deflection}
# ── Tokenise and compute TF-IDF for each speech within a section ─────────────

# We treat each section_id as a "document" and compute separate TF-IDF
# vectors for the question and the answer.

stop_words_oir <- c(
  tidytext::stop_words$word,
  "minister", "question", "dáil", "ask", "asked", "will", "can",
  "government", "department", "ireland", "irish"
)

tfidf_questions <- qa_pairs |>
  select(section_id, text = question_text) |>
  unnest_tokens(word, text) |>
  filter(!word %in% stop_words_oir, nchar(word) > 2) |>
  count(section_id, word) |>
  bind_tf_idf(word, section_id, n)

tfidf_answers <- qa_pairs |>
  select(section_id, text = answer_text) |>
  unnest_tokens(word, text) |>
  filter(!word %in% stop_words_oir, nchar(word) > 2) |>
  count(section_id, word) |>
  bind_tf_idf(word, section_id, n)

# ── Compute cosine similarity per section_id ──────────────────────────────────
section_ids <- intersect(
  unique(tfidf_questions$section_id),
  unique(tfidf_answers$section_id)
)

deflection_scores <- map_dfr(section_ids, function(sid) {
  q_vec <- tfidf_questions |>
    filter(section_id == sid) |>
    select(word, tf_idf) |>
    tibble::deframe()

  a_vec <- tfidf_answers |>
    filter(section_id == sid) |>
    select(word, tf_idf) |>
    tibble::deframe()

  tibble(
    section_id  = sid,
    cosine_sim  = cosine_sim(q_vec, a_vec)
  )
})

# ── Attach to main pairs table ────────────────────────────────────────────────
qa_pairs <- qa_pairs |>
  left_join(deflection_scores, by = "section_id")

# ── Distribution of deflection scores ────────────────────────────────────────
ggplot(qa_pairs, aes(x = cosine_sim)) +
  geom_histogram(bins = 40, fill = "#AB47BC", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = median(qa_pairs$cosine_sim, na.rm = TRUE),
             linetype = "dashed", colour = "#6A1B9A", linewidth = 0.8) +
  annotate("text",
           x     = median(qa_pairs$cosine_sim, na.rm = TRUE) + 0.02,
           y     = Inf, vjust = 1.5, hjust = 0,
           label = sprintf("Median = %.2f",
                           median(qa_pairs$cosine_sim, na.rm = TRUE)),
           colour = "#6A1B9A", size = 3.5) +
  scale_x_continuous(limits = c(0, 1), labels = number_format(accuracy = 0.01)) +
  labs(
    title    = "Distribution of Deflection Scores (Cosine Similarity)",
    subtitle = "Higher = answer vocabulary aligns with question; lower = deflection",
    x        = "Cosine similarity (question TF-IDF vs answer TF-IDF)",
    y        = "Number of Q&A exchanges",
    caption  = "Source: Oireachtas Open Data API via oiReachtas"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold")
  )
```

---

## 6  Readability: Flesch-Kincaid Score

The **Flesch-Kincaid Reading Ease** formula rates text on a 0–100 scale;
higher scores mean the text is easier to read. We implement it directly in
R without additional packages.

```
  FK = 206.835 − 1.015 × (words / sentences) − 84.6 × (syllables / words)
```

Syllables are approximated by counting contiguous vowel groups in each word
(a robust heuristic that works well for parliamentary English).

```{r readability-helper}
# ── Syllable counter (vowel-group heuristic) ──────────────────────────────────
count_syllables <- function(word) {
  word  <- tolower(trimws(word))
  # Remove trailing silent 'e' before counting
  word  <- sub("e$", "", word)
  # Count vowel groups
  groups <- stringr::str_count(word, "[aeiouy]+")
  # Every word has at least one syllable
  pmax(groups, 1L)
}

# ── Flesch-Kincaid Reading Ease ───────────────────────────────────────────────
#  Input:  character vector of one or more paragraphs / speeches
#  Output: scalar FK score

flesch_kincaid <- function(text) {
  text <- paste(text, collapse = " ")

  # ── Sentence count: split on .  !  ? ──────────────────────────────────────
  sentences <- stringr::str_split(text, "[.!?]+")[[1]]
  sentences <- sentences[nchar(trimws(sentences)) > 0]
  n_sent    <- max(length(sentences), 1L)

  # ── Word tokens ────────────────────────────────────────────────────────────
  words     <- unlist(stringr::str_split(text, "\\s+"))
  words     <- words[nchar(trimws(words)) > 0]
  # Strip punctuation from each token before syllable counting
  words     <- stringr::str_replace_all(words, "[^a-zA-Z]", "")
  words     <- words[nchar(words) > 0]
  n_words   <- max(length(words), 1L)

  # ── Syllable count ─────────────────────────────────────────────────────────
  n_syll    <- sum(count_syllables(words))

  # ── FK formula ─────────────────────────────────────────────────────────────
  206.835 - 1.015 * (n_words / n_sent) - 84.6 * (n_syll / n_words)
}
```

```{r apply-readability}
# ── Apply FK score to every answer ───────────────────────────────────────────
qa_pairs <- qa_pairs |>
  mutate(
    fk_score = map_dbl(answer_text, flesch_kincaid)
  )

# Clamp to [0, 100] — outliers from very short or malformed text
qa_pairs <- qa_pairs |>
  mutate(fk_score = pmax(0, pmin(100, fk_score)))

# ── Mean readability per minister ─────────────────────────────────────────────
readability_by_minister <- qa_pairs |>
  filter(!is.na(minister_name)) |>
  group_by(minister_name, department) |>
  summarise(
    mean_fk      = mean(fk_score,    na.rm = TRUE),
    median_fk    = median(fk_score,  na.rm = TRUE),
    n_answers    = n(),
    .groups      = "drop"
  ) |>
  filter(n_answers >= 5) |>          # require at least 5 exchanges for stability
  arrange(desc(mean_fk))

# ── Lollipop: top and bottom 10 by readability ────────────────────────────────
readability_plot <- bind_rows(
  slice_max(readability_by_minister, mean_fk, n = 10),
  slice_min(readability_by_minister, mean_fk, n = 10)
) |>
  distinct() |>
  mutate(
    minister_name = forcats::fct_reorder(minister_name, mean_fk),
    label_pos     = ifelse(mean_fk > 50, mean_fk - 1.5, mean_fk + 1.5),
    hjust_val     = ifelse(mean_fk > 50, 1, 0)
  )

ggplot(readability_plot,
       aes(x = mean_fk, y = minister_name, colour = mean_fk)) +
  geom_segment(aes(x = 0, xend = mean_fk, yend = minister_name),
               linewidth = 0.6, colour = "grey80") +
  geom_point(size = 3.5) +
  scale_colour_gradient(low = "#EF5350", high = "#26A69A",
                        name = "FK score", limits = c(0, 100)) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  labs(
    title    = "Ministerial Answer Readability (Flesch-Kincaid)",
    subtitle = "Higher score = easier to read · top and bottom 10 ministers shown",
    x        = "Mean Flesch-Kincaid Reading Ease score",
    y        = NULL,
    caption  = "Source: Oireachtas Open Data API via oiReachtas"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.title         = element_text(face = "bold"),
    legend.position    = "right"
  )
```

---

## 7  Combined Minister Scorecard: Bubble Chart

We now bring all three dimensions together. First we build a per-minister
summary, then we attach party labels from `get_members()` and render a
bubble chart where:

- **x-axis** = mean deflection score (cosine similarity)
- **y-axis** = mean readability (Flesch-Kincaid)
- **bubble size** = mean answer length (word count)
- **colour** = party affiliation

```{r scorecard-data}
# ── Per-minister summary ──────────────────────────────────────────────────────
minister_scorecard <- qa_pairs |>
  filter(!is.na(minister_name)) |>
  group_by(minister_name, department) |>
  summarise(
    mean_cosine    = mean(cosine_sim,   na.rm = TRUE),
    mean_fk        = mean(fk_score,     na.rm = TRUE),
    mean_words     = mean(answer_words, na.rm = TRUE),
    n_exchanges    = n(),
    .groups        = "drop"
  ) |>
  filter(n_exchanges >= 5)

# ── Attach party from member roster ──────────────────────────────────────────
members <- get_members(chamber = "dail", all_pages = TRUE) |>
  select(member_id = member_code, member_name = full_name) |>
  left_join(
    get_member_parties(chamber = "dail", all_pages = TRUE) |>
      group_by(member_id) |>
      slice_max(coalesce(end_date, Sys.Date()), n = 1) |>
      ungroup() |>
      select(member_id, party = party_code),
    by = "member_id"
  ) |>
  select(member_name, party)

minister_scorecard <- minister_scorecard |>
  left_join(members, by = c("minister_name" = "member_name")) |>
  mutate(party = coalesce(party, "Ind"))

# Map party codes to the canonical short forms used in party_colours
minister_scorecard <- minister_scorecard |>
  mutate(
    party_code = case_when(
      stringr::str_detect(party, regex("fianna|FF",  ignore_case = TRUE)) ~ "FF",
      stringr::str_detect(party, regex("fine gael|FG", ignore_case = TRUE)) ~ "FG",
      stringr::str_detect(party, regex("sinn féin|SF", ignore_case = TRUE)) ~ "SF",
      stringr::str_detect(party, regex("labour|Lab",   ignore_case = TRUE)) ~ "Lab",
      stringr::str_detect(party, regex("green|GP",     ignore_case = TRUE)) ~ "GP",
      stringr::str_detect(party, regex("social dem|SD", ignore_case = TRUE)) ~ "SD",
      stringr::str_detect(party, regex("people before|PBP", ignore_case = TRUE)) ~ "PBP",
      TRUE ~ "Ind"
    )
  )
```

```{r scorecard-plot}
# ── Bubble chart ──────────────────────────────────────────────────────────────
ggplot(minister_scorecard,
       aes(x      = mean_cosine,
           y      = mean_fk,
           size   = mean_words,
           colour = party_code,
           label  = minister_name)) +
  # Reference lines at medians
  geom_hline(yintercept = median(minister_scorecard$mean_fk,    na.rm = TRUE),
             linetype = "dashed", colour = "grey70", linewidth = 0.4) +
  geom_vline(xintercept = median(minister_scorecard$mean_cosine, na.rm = TRUE),
             linetype = "dashed", colour = "grey70", linewidth = 0.4) +
  # Quadrant annotations
  annotate("text", x = 0.02, y = 95, label = "Clear & deflecting",
           hjust = 0, vjust = 1, colour = "grey55", size = 3, fontface = "italic") +
  annotate("text", x = 0.98, y = 95, label = "Clear & responsive",
           hjust = 1, vjust = 1, colour = "grey55", size = 3, fontface = "italic") +
  annotate("text", x = 0.02, y = 5,  label = "Opaque & deflecting",
           hjust = 0, vjust = 0, colour = "grey55", size = 3, fontface = "italic") +
  annotate("text", x = 0.98, y = 5,  label = "Opaque & responsive",
           hjust = 1, vjust = 0, colour = "grey55", size = 3, fontface = "italic") +
  geom_point(alpha = 0.75) +
  geom_text_repel(size = 2.8, max.overlaps = 15,
                  segment.colour = "grey70", segment.size = 0.3,
                  show.legend = FALSE) +
  scale_colour_manual(values = party_colours, name = "Party",
                      drop = FALSE) +
  scale_size_continuous(range = c(3, 14), name = "Mean answer\nlength (words)",
                        labels = comma) +
  scale_x_continuous(limits = c(0, 1),
                     labels = number_format(accuracy = 0.01),
                     name   = "Mean deflection score (cosine similarity ↑ = more responsive)") +
  scale_y_continuous(limits = c(0, 100),
                     name   = "Mean readability (Flesch-Kincaid ↑ = clearer)") +
  labs(
    title    = "Ministerial Responsiveness Scorecard",
    subtitle = "Oral questions · Dáil Éireann · Jan–Mar 2024 · ministers with ≥ 5 exchanges",
    caption  = "Bubble size = mean answer word count · Source: Oireachtas Open Data API via oiReachtas"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold"),
    legend.position  = "right"
  )
```

---

## 8  Department-Level Heatmap

The bubble chart shows minister-level variation; the heatmap aggregates to
department level and normalises each metric to [0, 1] so the three dimensions
are directly comparable on a single scale.

```{r heatmap-data}
# ── Department-level summary ──────────────────────────────────────────────────
dept_summary <- qa_pairs |>
  filter(!is.na(department)) |>
  group_by(department) |>
  summarise(
    mean_cosine = mean(cosine_sim,   na.rm = TRUE),
    mean_fk     = mean(fk_score,     na.rm = TRUE),
    mean_words  = mean(answer_words, na.rm = TRUE),
    n_exchanges = n(),
    .groups     = "drop"
  ) |>
  filter(n_exchanges >= 10)           # at least 10 exchanges per department

# ── Min-max normalise each metric to [0, 1] ──────────────────────────────────
norm01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (diff(rng) == 0) return(rep(0.5, length(x)))
  (x - rng[1]) / diff(rng)
}

dept_normalised <- dept_summary |>
  mutate(
    deflection_norm  = norm01(mean_cosine),
    readability_norm = norm01(mean_fk),
    verbosity_norm   = norm01(mean_words)
  ) |>
  select(department,
         Deflection   = deflection_norm,
         Readability  = readability_norm,
         Verbosity    = verbosity_norm) |>
  pivot_longer(
    cols      = c(Deflection, Readability, Verbosity),
    names_to  = "metric",
    values_to = "score_norm"
  )

# Order departments by combined score (sum of normalised metrics)
dept_order_heat <- dept_normalised |>
  group_by(department) |>
  summarise(total = sum(score_norm), .groups = "drop") |>
  arrange(total) |>
  pull(department)

dept_normalised <- dept_normalised |>
  mutate(department = factor(department, levels = dept_order_heat))
```

```{r heatmap-plot}
ggplot(dept_normalised, aes(x = metric, y = department, fill = score_norm)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", score_norm)),
            size = 3, colour = "white", fontface = "bold") +
  scale_fill_gradient(
    low    = "#EF5350",
    high   = "#26A69A",
    name   = "Normalised\nscore (0–1)",
    limits = c(0, 1)
  ) +
  scale_x_discrete(position = "top") +
  labs(
    title    = "Department Responsiveness Heatmap",
    subtitle = "Metrics normalised 0–1 within each column · higher = better performance",
    x        = NULL,
    y        = NULL,
    caption  = "Source: Oireachtas Open Data API via oiReachtas"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x      = element_text(face = "bold", size = 11),
    panel.grid       = element_blank(),
    plot.title       = element_text(face = "bold"),
    legend.position  = "right"
  )
```

---

## 9  Extending to Written Questions via XML Parsing

The same methodology extends naturally to **written questions**, which have a
far larger volume (typically 10× more exchanges per week than oral questions).
Written question answers are not directly returned as `text` fields by
`get_debate_text()`; instead, the `formats_xml` column in `get_questions()`
output contains a URI to a structured XML document.

The general pattern for extracting written answer text from the XML is:

```{r written-xml, eval=FALSE}
# Written questions provide an XML URI in formats_xml
written_qs <- get_questions(
  date_start    = "2024-01-01",
  date_end      = "2024-03-31",
  question_type = "written",
  limit         = 1000
)

# Helper: fetch and parse a single written-question XML document
parse_written_xml <- function(xml_uri) {
  if (is.na(xml_uri) || xml_uri == "") return(NULL)

  tryCatch({
    raw_xml <- xml2::read_xml(xml_uri)

    # Written question XML structure (Oireachtas API):
    #   <questionList>
    #     <question questionId="..." questionType="written" date="...">
    #       <questionText> ... </questionText>
    #       <answer> ... </answer>
    #     </question>
    #   </questionList>

    questions_nodes <- xml2::xml_find_all(raw_xml, ".//question")

    purrr::map_dfr(questions_nodes, function(node) {
      tibble::tibble(
        question_id   = xml2::xml_attr(node, "questionId"),
        question_type = xml2::xml_attr(node, "questionType"),
        date          = xml2::xml_attr(node, "date"),
        question_text = xml2::xml_text(
          xml2::xml_find_first(node, ".//questionText")
        ),
        answer_text   = xml2::xml_text(
          xml2::xml_find_first(node, ".//answer")
        )
      )
    })
  }, error = function(e) {
    message(sprintf("Failed to parse XML at %s: %s", xml_uri, e$message))
    NULL
  })
}

# Map over all written question XML URIs
written_parsed <- written_qs |>
  dplyr::mutate(
    parsed = purrr::map(formats_xml, parse_written_xml)
  ) |>
  dplyr::filter(!purrr::map_lgl(parsed, is.null)) |>
  tidyr::unnest(parsed)

# From here, apply the same cosine_sim() and flesch_kincaid() functions
# defined earlier to written_parsed$question_text and
# written_parsed$answer_text to build a written-question scorecard.
#
# Written questions also expose the `department` column directly from
# get_questions(), so no regex extraction is needed.
```

> **Tip:** Written answers are frequently longer and more formulaic than oral
> answers; you may find that readability scores cluster more tightly and that
> deflection scores are systematically lower (ministers use boilerplate language
> that shares little vocabulary with the specific question). Stratifying the
> analysis by `question_type` ("oral" vs "written") before aggregating will
> surface these structural differences clearly.

---

## Summary

This vignette has demonstrated three complementary, fully reproducible measures
of ministerial quality in oral question time:

| Step | What we built |
|---|---|
| **4** | Violin plot of answer word counts by department (verbosity) |
| **5** | Cosine similarity between question and answer TF-IDF vectors (deflection) |
| **6** | Flesch-Kincaid reading ease computed from scratch in R (clarity) |
| **7** | Bubble chart combining all three metrics per minister, coloured by party |
| **8** | Department-level normalised heatmap for at-a-glance comparison |
| **9** | Pattern for extending the same pipeline to written questions via XML |

The three dimensions are intentionally independent: a minister can score well
on verbosity but poorly on deflection (long answer, different topic), or well
on readability but poorly on both others (clear but brief and off-topic). The
combined scorecard in Section 7 is the most informative single view — ministers
in the top-right quadrant (high cosine similarity, high readability) are both
responsive and accessible, which is the democratic ideal for oral question time.

```{r session-info, include=TRUE, eval=TRUE}
sessionInfo()
```
