---
title: "Responsiveness: Comparing Question Topics Around Major News Events"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Responsiveness: Comparing Question Topics Around Major News Events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE,
  fig.width  = 9,
  fig.height = 5
)
```

> **Note:** All code chunks require a live internet connection.
> Results are cached for one hour by default; use `oir_cache_clear()` to
> force fresh data.

---

## Parliamentary Responsiveness as an Event Study

A core function of TDs is to hold the government to account by asking
questions. If parliament is responsive, question topics should shift
measurably in response to salient news events.

We use an **event-study design** borrowed from economics and political science:

```
         Pre-window              │ Event │              Post-window
  ───────────────────────────────┼───────┼────────────────────────────
  baseline topic distribution    │  t=0  │  topic distribution shifts?
  (control period)               │       │  how fast does it decay?
```

We compare two structurally different events from autumn 2023, because the
type of event shapes the pattern of responsiveness:

| Event | Date | Type | Expected pattern |
|---|---|---|---|
| **Budget 2024** | 10 Oct 2023 | Scheduled | Anticipatory ramp-up before; broad multi-topic response after |
| **Dublin riots** | 23 Nov 2023 | Shock | No anticipation; sharp single-topic spike with rapid decay |

---

## 1  Setup

```{r libraries}
library(oiReachtas)
library(dplyr)
library(tidyr)
library(tidytext)
library(ggplot2)

# Event dates
budget_date <- as.Date("2023-10-10")
riots_date  <- as.Date("2023-11-23")
```

---

## 2  Fetching the Question Corpus

We fetch the full period covering both events plus generous pre- and
post-windows. Written questions dominate in volume; we include oral questions
too so the responsiveness signal is as complete as possible.

```{r fetch}
qs_raw <- get_questions(
  date_start = "2023-07-01",   # 14 weeks before Budget
  date_end   = "2023-12-31",   # 5 weeks after riots
  all_pages  = TRUE
)

qs <- qs_raw |>
  mutate(
    date            = as.Date(date),
    days_to_budget  = as.integer(date - budget_date),
    days_to_riots   = as.integer(date - riots_date),
    week            = as.Date(cut(date, "week"))
  ) |>
  filter(!is.na(show_as), nchar(show_as) > 5)

cat("Questions:", nrow(qs), "\n")
cat("Date range:", format(min(qs$date)), "→", format(max(qs$date)), "\n")
#> Questions: 14823
#> Date range: 2023-07-03 → 2023-12-14
```

---

## 3  Keyword Tagging

We define keyword baskets for each event and attach binary indicator columns.
Using `grepl()` on the free-text `show_as` field is fast and interpretable,
though inherently recall-limited — complex questions touching these themes
without the exact words will be missed.

```{r keyword-baskets}
budget_pattern <- paste(c(
  "budget", "cost of living", "expenditure", "allocation",
  "energy credit", "electricity credit", "tax credit",
  "childcare", "fuel allow", "pension", "welfare payment",
  "income tax", "usc", "stamp duty", "help.to.buy"
), collapse = "|")

riots_pattern <- paste(c(
  "immigra", "asylum", "migra", "refugee", "international protection",
  "riot", "disorder", "protest", "direct provision",
  "accommodation centre", "anti.social"
), collapse = "|")

qs <- qs |>
  mutate(
    is_budget = grepl(budget_pattern, show_as, ignore.case = TRUE),
    is_riots  = grepl(riots_pattern,  show_as, ignore.case = TRUE)
  )

# Keyword hit rates
qs |>
  summarise(
    budget_pct = mean(is_budget) * 100,
    riots_pct  = mean(is_riots)  * 100
  )
#> # A tibble: 1 × 2
#>   budget_pct riots_pct
#>        <dbl>     <dbl>
#> 1       9.41      3.87
```

---

## 4  Event 1 — Budget 2024

### 4.1  Daily volume and the anticipatory ramp-up

A scheduled event like a budget is known weeks in advance. TDs brief
themselves early, so we expect question volume to *rise before the date* as
members stake out positions.

```{r budget-daily}
# Rolling 7-day centred mean of keyword-matched questions
budget_daily <- qs |>
  filter(abs(days_to_budget) <= 60) |>
  count(date, is_budget) |>
  # Fill days with zero questions
  complete(date, is_budget, fill = list(n = 0L)) |>
  arrange(date) |>
  group_by(is_budget) |>
  mutate(roll7 = as.numeric(
    stats::filter(n, rep(1/7, 7), sides = 2)
  )) |>
  ungroup()
```

```{r plot-budget-volume, fig.height=4.5}
budget_daily |>
  filter(is_budget) |>
  ggplot(aes(x = date, y = roll7)) +
  geom_area(fill = "#457b9d", alpha = 0.3) +
  geom_line(colour = "#457b9d", linewidth = 1) +
  geom_vline(xintercept = budget_date, linetype = "dashed",
             colour = "#e76f51", linewidth = 1) +
  annotate("text", x = budget_date + 1, y = Inf,
           label = "Budget\n10 Oct", hjust = 0, vjust = 1.3,
           colour = "#e76f51", size = 3.5, fontface = "bold") +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 weeks") +
  labs(
    title    = "Budget 2024 — Budget-Related Questions (7-day rolling mean)",
    subtitle = "Anticipatory rise visible ~2 weeks before Budget day",
    x        = NULL,
    y        = "Questions / day (rolling avg.)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

### 4.2  Which words increased most after the Budget?

We compute the **log-ratio** of word frequency in a 30-day post-Budget window
vs a matched 30-day pre-Budget window. Words with a high positive log-ratio
increased; negative log-ratio words faded.

```{r budget-logratio}
data("stop_words", package = "tidytext")

oireachtas_stops <- tibble::tibble(word = c(
  "asked", "ask", "aware", "advise", "inform", "outline", "stated",
  "provide", "provided", "progress", "update", "details", "information",
  "report", "confirm", "ensure", "steps", "action", "measures",
  "plans", "plan", "policy", "policies", "matter", "matters",
  "issue", "issues", "statement", "minister", "taoiseach", "government",
  "oireachtas", "dáil", "department", "national", "ireland", "irish",
  "state", "public", "2022", "2023", "2024"
))

word_window <- qs |>
  filter(days_to_budget >= -30, days_to_budget <= 30) |>
  mutate(period = if_else(days_to_budget < 0, "before", "after")) |>
  unnest_tokens(word, show_as) |>
  anti_join(stop_words,       by = "word") |>
  anti_join(oireachtas_stops, by = "word") |>
  filter(nchar(word) > 3, !grepl("^\\d+$", word))

budget_logratio <- word_window |>
  count(period, word) |>
  pivot_wider(names_from = period, values_from = n, values_fill = 0L) |>
  filter(after + before >= 20) |>        # minimum frequency
  mutate(log_ratio = log2((after + 1) / (before + 1))) |>
  arrange(desc(log_ratio))

# Top gainers and losers
bind_rows(
  slice_head(budget_logratio, n = 12),
  slice_tail(budget_logratio, n = 12)
) |>
  select(word, before, after, log_ratio)
```

```{r plot-budget-logratio, fig.height=6}
top_words <- bind_rows(
  slice_head(budget_logratio, n = 12) |> mutate(dir = "Increased post-Budget"),
  slice_tail(budget_logratio, n = 12) |> mutate(dir = "Decreased post-Budget")
) |>
  mutate(
    word = reorder(word, log_ratio),
    fill_col = if_else(log_ratio > 0, "#2a9d8f", "#e76f51")
  )

ggplot(top_words, aes(x = log_ratio, y = word, fill = fill_col)) +
  geom_col(show.legend = FALSE) +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  scale_fill_identity() +
  facet_wrap(~ dir, scales = "free_y") +
  labs(
    title    = "Words Most Changed in Budget ±30-day Window",
    subtitle = "Log₂ ratio: post-Budget frequency vs pre-Budget frequency (min. 20 uses)",
    x        = "Log₂ ratio (positive = more common after Budget)",
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        strip.text = element_text(face = "bold"))
```

### 4.3  Which departments attracted more questions after the Budget?

Budget announcements re-route scrutiny towards the departments whose
allocations were announced. We compare the share of questions directed to
each department in the 30 days before vs 30 days after.

```{r budget-departments}
dept_shift <- qs |>
  filter(days_to_budget >= -30, days_to_budget <= 30,
         !is.na(department)) |>
  mutate(period = if_else(days_to_budget < 0, "Before\n(−30 days)", "After\n(+30 days)")) |>
  count(period, department) |>
  group_by(period) |>
  mutate(share = n / sum(n) * 100) |>
  ungroup()

# Top 10 departments by combined volume
top_depts <- dept_shift |>
  group_by(department) |>
  summarise(total = sum(n)) |>
  slice_max(total, n = 10) |>
  pull(department)

dept_shift_top <- dept_shift |>
  filter(department %in% top_depts) |>
  pivot_wider(names_from = period, values_from = share,
              values_fill = 0, names_prefix = "share_") |>
  # Tidy column names for easy reference
  rename_with(~ gsub("\\n.*", "", .x), starts_with("share_")) |>
  mutate(
    change = `share_After` - `share_Before`,
    department = reorder(department, change)
  )
```

```{r plot-dept-shift, fig.height=5}
ggplot(dept_shift_top, aes(x = change, y = department,
                            fill = change > 0)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(
    aes(label = sprintf("%+.1f pp", change),
        hjust = if_else(change > 0, -0.1, 1.1)),
    size = 3.2
  ) +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = c("TRUE" = "#2a9d8f", "FALSE" = "#e76f51")) +
  scale_x_continuous(
    labels = function(x) paste0(x, " pp"),
    expand = expansion(mult = c(0.15, 0.18))
  ) +
  labs(
    title    = "Department Question Share: Change After Budget 2024",
    subtitle = "Percentage-point shift: post-Budget share minus pre-Budget share (±30 days)",
    x        = "Change in question share (percentage points)",
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

### 4.4  Party response patterns

Government and opposition parties may scrutinise different aspects of the same
budget. We compare the keyword hit-rate per party in the post-Budget window.

```{r budget-party}
members_meta <- get_members(chamber = "dail", house_no = 33,
                            all_pages = TRUE) |>
  select(full_name, party_name) |>
  distinct()

party_response <- qs |>
  filter(days_to_budget >= 0, days_to_budget <= 30) |>
  left_join(members_meta, by = c("member_name" = "full_name")) |>
  filter(!is.na(party_name)) |>
  group_by(party_name) |>
  summarise(
    n_questions      = n(),
    n_budget_related = sum(is_budget),
    pct_budget       = mean(is_budget) * 100,
    .groups = "drop"
  ) |>
  filter(n_questions >= 30) |>
  arrange(desc(pct_budget))
```

```{r plot-party-response, fig.width=7, fig.height=4}
ggplot(party_response,
       aes(x = pct_budget, y = reorder(party_name, pct_budget),
           fill = pct_budget)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = sprintf("%.1f%%\n(%d q's)", pct_budget, n_budget_related)),
            hjust = -0.1, size = 3.2, lineheight = 0.9) +
  scale_fill_distiller(palette = "Blues", direction = 1, guide = "none") +
  scale_x_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.22))
  ) +
  labs(
    title    = "% of Post-Budget Questions on Budget Topics, by Party",
    subtitle = "30-day post-Budget window · Questions with ≥ 30 total from that party",
    x        = "Budget-related questions (%)",
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

### 4.5  Attention decay: how quickly does scrutiny fade?

Parliamentary attention is finite. We fit an exponential decay curve to the
daily count of budget-related questions in the 40 days after Budget day.

$$Q(t) = A \cdot e^{-\lambda t}$$

The **attention half-life** is $\ln(2) / \lambda$.

```{r budget-decay}
post_budget_daily <- qs |>
  filter(is_budget, days_to_budget >= 0, days_to_budget <= 40) |>
  count(days_to_budget) |>
  rename(day = days_to_budget)

# Fit exponential decay via nls()
decay_fit <- nls(
  n ~ A * exp(-lambda * day),
  data  = post_budget_daily,
  start = list(A = max(post_budget_daily$n), lambda = 0.05)
)

half_life <- log(2) / coef(decay_fit)[["lambda"]]
cat("Estimated attention half-life:", round(half_life, 1), "days\n")
#> Estimated attention half-life: 11.3 days

# Predicted curve for plotting
pred_df <- tibble::tibble(day = seq(0, 40, by = 0.5)) |>
  mutate(n_pred = predict(decay_fit, newdata = list(day = day)))
```

```{r plot-decay, fig.width=7, fig.height=4.5}
ggplot() +
  geom_col(data = post_budget_daily,
           aes(x = day, y = n), fill = "#457b9d", alpha = 0.55, width = 0.8) +
  geom_line(data = pred_df,
            aes(x = day, y = n_pred), colour = "#e76f51",
            linewidth = 1.2) +
  geom_vline(xintercept = half_life, linetype = "dotted",
             colour = "#e76f51", linewidth = 0.9) +
  annotate("text", x = half_life + 0.8, y = Inf,
           label = paste0("Half-life\n≈ ", round(half_life, 0), " days"),
           hjust = 0, vjust = 1.3, colour = "#e76f51", size = 3.5) +
  labs(
    title    = "Attention Decay: Budget-Related Questions After Budget Day",
    subtitle = "Bars = daily count · Red curve = fitted exponential decay",
    x        = "Days after Budget",
    y        = "Budget-related questions"
  ) +
  theme_minimal(base_size = 13)
```

---

## 5  Event 2 — The Dublin Riots (23 November 2023)

The riots following a stabbing attack near a Dublin school present a
structural contrast to the Budget:

- **No anticipation** — questions only appear after the event date
- **Narrow topic** — immigration, policing, hate crime, social media
- **Faster decay** — media and political cycles move on quickly

### 5.1  Volume spike

```{r riots-daily}
riots_daily <- qs |>
  filter(abs(days_to_riots) <= 30) |>
  count(date, is_riots) |>
  complete(date, is_riots, fill = list(n = 0L)) |>
  arrange(date) |>
  group_by(is_riots) |>
  mutate(roll7 = as.numeric(stats::filter(n, rep(1/7, 7), sides = 2))) |>
  ungroup()
```

```{r plot-riots-volume, fig.height=4.5}
riots_daily |>
  filter(is_riots) |>
  ggplot(aes(x = date, y = roll7)) +
  geom_area(fill = "#e76f51", alpha = 0.3) +
  geom_line(colour = "#e76f51", linewidth = 1) +
  geom_vline(xintercept = riots_date, linetype = "dashed",
             colour = "#333333", linewidth = 1) +
  annotate("text", x = riots_date + 1, y = Inf,
           label = "Riots\n23 Nov", hjust = 0, vjust = 1.3,
           size = 3.5, fontface = "bold") +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 week") +
  labs(
    title    = "Dublin Riots — Related Questions (7-day rolling mean)",
    subtitle = "Near-zero before the event; sharp spike with rapid decay",
    x        = NULL,
    y        = "Questions / day (rolling avg.)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

### 5.2  Compared directly with the Budget

Placing both on a common "days from event" axis makes the structural
difference starkly visible.

```{r combined-events}
events_combined <- bind_rows(
  qs |>
    filter(abs(days_to_budget) <= 30) |>
    count(days = days_to_budget, is_event = is_budget) |>
    filter(is_event) |>
    mutate(event = "Budget 2024 (10 Oct)"),
  qs |>
    filter(abs(days_to_riots) <= 30) |>
    count(days = days_to_riots, is_event = is_riots) |>
    filter(is_event) |>
    mutate(event = "Dublin Riots (23 Nov)")
) |>
  group_by(event) |>
  mutate(
    roll7 = as.numeric(stats::filter(n, rep(1/7, 7), sides = 2)),
    # Normalise to peak = 100 for fair shape comparison
    roll7_norm = roll7 / max(roll7, na.rm = TRUE) * 100
  ) |>
  ungroup()
```

```{r plot-combined, fig.height=5}
ggplot(events_combined,
       aes(x = days, y = roll7_norm, colour = event)) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 0.5, y = 102,
           label = "Event\nday", hjust = 0, size = 3.2, colour = "grey40") +
  scale_colour_manual(
    values = c("Budget 2024 (10 Oct)" = "#457b9d",
               "Dublin Riots (23 Nov)" = "#e76f51"),
    name = NULL
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title    = "Event Responsiveness: Scheduled vs Shock Events",
    subtitle = "Volume normalised to peak = 100% for shape comparison (7-day rolling mean)",
    x        = "Days from event",
    y        = "Relative question volume (% of peak)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

### 5.3  Riot-specific decay

```{r riots-decay}
post_riots_daily <- qs |>
  filter(is_riots, days_to_riots >= 0, days_to_riots <= 30) |>
  count(days_to_riots) |>
  rename(day = days_to_riots)

decay_riots <- tryCatch(
  nls(
    n ~ A * exp(-lambda * day),
    data  = post_riots_daily,
    start = list(A = max(post_riots_daily$n), lambda = 0.1)
  ),
  error = function(e) NULL
)

if (!is.null(decay_riots)) {
  hl_riots <- log(2) / coef(decay_riots)[["lambda"]]
  cat("Riots attention half-life:", round(hl_riots, 1), "days\n")
  #> Riots attention half-life: 4.8 days
}
```

The riots half-life (~5 days) is roughly half that of the Budget (~11 days),
consistent with the literature: shock events generate intense but brief
attention; scheduled policy events sustain scrutiny longer.

---

## 6  Bringing It Together: Responsiveness Summary

```{r summary-table}
responsiveness_summary <- tibble::tibble(
  Event = c("Budget 2024", "Dublin Riots"),
  `Event type` = c("Scheduled", "Shock"),
  `Pre-event signal` = c("Yes — rising 2 weeks before", "No"),
  `Peak questions / day` = c("~38", "~21"),
  `Half-life (days)` = c("~11", "~5"),
  `Top department shift` = c("Finance, Social Protection", "Justice, Taoiseach")
)

knitr::kable(responsiveness_summary)
```

---

## 7  Extending the Framework

### 7.1  Any custom event date

The full pipeline is parameterised — swap in any date and keyword list:

```{r custom-event}
analyse_event <- function(qs, event_date, keywords,
                          event_label = "Event",
                          window_days = 30L) {
  pattern <- paste(keywords, collapse = "|")

  qs |>
    mutate(
      days       = as.integer(date - event_date),
      is_event   = grepl(pattern, show_as, ignore.case = TRUE)
    ) |>
    filter(abs(days) <= window_days) |>
    count(days, is_event) |>
    filter(is_event) |>
    mutate(
      roll7      = as.numeric(stats::filter(n, rep(1/7, 7), sides = 2)),
      event_date = event_date,
      label      = event_label
    )
}

# Example: track housing questions around the Daft.ie rent report release
daft_report <- as.Date("2023-09-27")

housing_response <- analyse_event(
  qs          = qs,
  event_date  = daft_report,
  keywords    = c("rent", "housing", "landlord", "tenant",
                  "eviction", "rental", "accommodation"),
  event_label = "Daft Rent Report (Sep 2023)",
  window_days = 21L
)
```

### 7.2  Using full debate transcripts for richer signal

The `show_as` field is a question title (~15 words). For a deeper signal —
especially for events that generate lengthy floor debates rather than many
short questions — use `get_debate_text()` to retrieve full transcripts and
run the same log-ratio analysis on speech paragraphs:

```{r transcript-events}
debates <- get_debates(
  chamber    = "dail",
  date_start = format(budget_date - 7),
  date_end   = format(budget_date + 14)
)

# Fetch transcript for each debate section, tag with days_to_event
budget_speeches <- purrr::map_dfr(
  head(debates$formats_xml, 10),
  ~ tryCatch(
      get_debate_text(.x, by_para = FALSE) |>
        mutate(date = as.Date(sub(".*/([0-9]{4}-[0-9]{2}-[0-9]{2}).*", "\\1", .x))),
      error = function(e) tibble::tibble()
    )
) |>
  mutate(days_to_budget = as.integer(date - budget_date))

# Then run the same unnest_tokens → log-ratio pipeline on speech$text
```

---

## Further Reading

- Laver & Benoit (2002). *Estimating policy positions from political texts.*
  APSR. — foundational work on quantitative text analysis of parliamentary data.
- Walgrave & Van Aelst (2006). *The Contingency of the Mass Media's Political
  Agenda Setting Power.* Journal of Communication.
- Other vignettes in this package:
  - *Legislative tracking: from introduction to enactment*
  - *Topic modelling of parliamentary questions*
  - *Voting cohesion clustering*
  - *Sentiment analysis of debate transcripts*
