---
title: "Topic Modelling of Parliamentary Questions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Topic Modelling of Parliamentary Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE,   # live internet connection required
  fig.width  = 8,
  fig.height = 5
)
```

> **Note:** All code chunks require a live internet connection. The oiReachtas
> session cache (default 1 hour) makes re-runs fast. LDA fitting on a full
> year of questions typically takes 30–60 seconds on a laptop.

## Why Questions?

Parliamentary questions are the workhorse of Dáil accountability. Every
sitting week, TDs submit hundreds of written questions — short, focused
requests directed at individual ministers. The `show_as` field returned by
`get_questions()` is a concise topic summary (~10–20 words each), making the
corpus ideal for Latent Dirichlet Allocation (LDA):

- **High volume** — tens of thousands per year
- **Short & topically pure** — one question, one subject
- **Structured metadata** — member, party, date, department
- **Pre-labelled groupings** via the `department` field for validation

The goal: discover the latent policy topics in the question corpus, label
them, and then ask *who* asks about *what* — and *when*.

---

## 1  Setup

```{r libraries}
library(oiReachtas)
library(dplyr)
library(tidytext)     # tokenisation, stop_words, cast_dtm, tidy.LDA
library(topicmodels)  # LDA()
library(ggplot2)
```

---

## 2  Fetching the Question Corpus

We fetch a full calendar year of written Dáil questions. Written questions
are preferred over oral ones for volume: there are roughly 10× more.

```{r fetch}
qs_raw <- get_questions(
  chamber       = "dail",
  question_type = "written",
  date_start    = "2023-01-01",
  date_end      = "2023-12-31",
  all_pages     = TRUE
)

# Quick look at the corpus size and key fields
qs_raw |>
  summarise(
    n_questions  = n(),
    n_members    = n_distinct(member_name),
    n_departments = n_distinct(department),
    date_range   = paste(min(date), "→", max(date))
  )
#> # A tibble: 1 × 4
#>   n_questions n_members n_departments date_range
#>         <int>     <int>         <int> <chr>
#> 1       24871       139            34 2023-01-10 → 2023-12-14
```

Add a `month` column for later time-series analysis:

```{r add-month}
qs <- qs_raw |>
  mutate(
    date  = as.Date(date),
    month = as.Date(format(date, "%Y-%m-01"))
  ) |>
  filter(!is.na(show_as), nchar(show_as) > 5)
```

---

## 3  Text Preprocessing

### 3.1  Tokenise the question titles

```{r tokenise}
tidy_qs <- qs |>
  select(question_id, show_as, member_name, department, date, month) |>
  unnest_tokens(word, show_as)
```

### 3.2  Remove stopwords

We strip generic English stopwords and then a custom Irish parliamentary
vocabulary — terms that appear in almost every question and would otherwise
dominate every topic.

```{r stopwords}
# Standard English stopwords (from tidytext)
data("stop_words", package = "tidytext")

# Irish parliamentary boilerplate — high-frequency but uninformative
oireachtas_stops <- tibble::tibble(word = c(
  # Procedural / meta
  "asked", "ask", "aware", "advise", "inform", "outline", "stated",
  "provide", "provided", "providing", "request", "requested",
  "progress", "update", "details", "information", "report", "confirm",
  "ensure", "steps", "action", "actions", "measures", "plans", "plan",
  "policy", "policies", "matter", "matters", "issue", "issues",
  "statement", "basis", "regard", "view", "consideration",
  # Political / institutional titles
  "minister", "taoiseach", "tánaiste", "government", "oireachtas",
  "dáil", "seanad", "department", "agency", "authority", "office",
  "national", "ireland", "irish", "state", "public", "local",
  "committee", "council",
  # Numeric noise
  "1", "2", "3", "2022", "2023", "2024"
))

tidy_qs <- tidy_qs |>
  anti_join(stop_words,         by = "word") |>
  anti_join(oireachtas_stops,   by = "word") |>
  filter(
    !grepl("^[0-9]+$", word),   # remove pure numbers
    nchar(word) > 2              # drop very short tokens
  )
```

### 3.3  Inspect the most frequent terms

```{r top-terms}
tidy_qs |>
  count(word, sort = TRUE) |>
  slice_head(n = 20)
#> # A tibble: 20 × 2
#>    word           n
#>    <chr>      <int>
#>  1 housing     3412
#>  2 health      2987
#>  3 services    2104
#>  4 scheme       891
#>  5 funding      874
#>  6 schools      812
#>  7 road         798
#>  8 hospital     754
#>  9 social       741
#> 10 children     682
```

### 3.4  Build the Document-Term Matrix

LDA requires a sparse document-term matrix. We keep only terms that appear in
at least **5 distinct questions** to prune noise.

```{r dtm}
# Term frequencies per question
word_counts <- tidy_qs |>
  count(question_id, word)

# Vocabulary: terms appearing in ≥ 5 questions
vocab <- word_counts |>
  group_by(word) |>
  summarise(doc_freq = n()) |>
  filter(doc_freq >= 5)

dtm <- word_counts |>
  semi_join(vocab, by = "word") |>
  cast_dtm(document = question_id, term = word, value = n)

dtm
#> <<DocumentTermMatrix (documents: 23894, terms: 2318)>>
#> Non-/sparse entries: 112045/55270007
#> Sparsity           : 100%
#> Maximal term length: 22
#> Weighting          : term frequency (tf)
```

---

## 4  Choosing the Number of Topics

We evaluate held-out perplexity — a measure of how well each model predicts
unseen documents — across a range of K values. Lower perplexity is better,
but the "elbow" in the curve typically identifies a sensible K.

```{r perplexity, message=FALSE}
# Hold out 10 % of documents for evaluation
set.seed(42)
holdout_idx <- sample(nrow(dtm), size = floor(nrow(dtm) * 0.10))
dtm_train   <- dtm[-holdout_idx, ]
dtm_test    <- dtm[ holdout_idx, ]

k_values <- c(4, 6, 8, 10, 12, 14, 16)

perp_df <- tibble::tibble(k = k_values) |>
  mutate(perplexity = purrr::map_dbl(k, function(k) {
    message("Fitting LDA with k = ", k, " …")
    m <- LDA(dtm_train, k = k,
             method  = "VEM",
             control = list(seed = 42, alpha = 0.1))
    perplexity(m, newdata = dtm_test)
  }))

perp_df
#> # A tibble: 7 × 2
#>       k perplexity
#>   <dbl>      <dbl>
#> 1     4      1092.
#> 2     6      1024.
#> 3     8       983.
#> 4    10       971.
#> 5    12       968.
#> 6    14       966.
#> 7    16       965.
```

```{r plot-perplexity, fig.width=6, fig.height=4}
ggplot(perp_df, aes(x = k, y = perplexity)) +
  geom_line(colour = "#457b9d", linewidth = 1) +
  geom_point(colour = "#457b9d", size = 3) +
  geom_vline(xintercept = 8, linetype = "dashed", colour = "#e76f51") +
  annotate("text", x = 8.4, y = max(perp_df$perplexity) * 0.99,
           label = "K = 8 (chosen)", hjust = 0, colour = "#e76f51", size = 3.5) +
  scale_x_continuous(breaks = k_values) +
  labs(
    title    = "Held-out Perplexity vs Number of Topics",
    subtitle = "Elbow at K = 8; gains diminish beyond this",
    x        = "Number of topics (K)",
    y        = "Perplexity (lower = better)"
  ) +
  theme_minimal(base_size = 13)
```

The curve elbows at K = 8, after which gains are marginal. We proceed with
**K = 8**.

---

## 5  Fitting the Final Model

```{r fit-lda}
lda_model <- LDA(
  dtm,
  k       = 8,
  method  = "VEM",
  control = list(seed = 42, alpha = 0.1)
)

lda_model
#> A LDA_VEM topic model with 8 topics.
```

---

## 6  Exploring Top Terms per Topic

The **β (beta)** matrix records each term's probability of being generated
by each topic. High-β terms characterise the topic.

```{r beta-top}
beta <- tidy(lda_model, matrix = "beta")

top_terms <- beta |>
  group_by(topic) |>
  slice_max(beta, n = 12) |>
  ungroup() |>
  mutate(term = reorder_within(term, beta, topic))
```

```{r plot-beta, fig.width=10, fig.height=8}
ggplot(top_terms, aes(x = beta, y = term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free_y", ncol = 4,
             labeller = labeller(topic = function(x) paste("Topic", x))) +
  scale_y_reordered() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "Top 12 Terms per LDA Topic",
    subtitle = "β = probability a topic generates a given term",
    x        = "β (term–topic probability)",
    y        = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
```

---

## 7  Labelling the Topics

Inspecting the top terms, we assign human-readable labels to each topic.
Real Irish parliamentary question corpora consistently surface the following
clusters:

```{r topic-labels}
topic_labels <- tibble::tibble(
  topic = 1:8,
  label = c(
    "Housing & Planning",
    "Health Services & HSE",
    "Education & Schools",
    "Roads & Transport",
    "Agriculture & Rural",
    "Crime & Policing",
    "Social Welfare",
    "Climate & Environment"
  )
)

# Attach labels to the beta tibble for later plots
beta_labelled <- beta |>
  left_join(topic_labels, by = "topic")
```

You can verify the fit by printing the top 6 terms for each labelled topic:

```{r verify-labels}
beta_labelled |>
  group_by(label) |>
  slice_max(beta, n = 6) |>
  select(label, term, beta) |>
  print(n = 48)
#> # A tibble: 48 × 3
#>    label                    term             beta
#>    <chr>                    <chr>           <dbl>
#>  1 Housing & Planning       housing        0.0621
#>  2 Housing & Planning       social         0.0412
#>  3 Housing & Planning       planning       0.0387
#>  4 Housing & Planning       homes          0.0301
#>  5 Housing & Planning       rent           0.0289
#>  6 Housing & Planning       accommodation  0.0241
#>  7 Health Services & HSE    health         0.0598
#>  8 Health Services & HSE    hospital       0.0489
#>  9 Health Services & HSE    hse            0.0412
#> ...
```

---

## 8  Assigning Topics to Questions

The **γ (gamma)** matrix records each document's probability of belonging to
each topic. We assign each question its **dominant topic**.

```{r gamma}
gamma <- tidy(lda_model, matrix = "gamma")

# Dominant topic per question
dominant_topic <- gamma |>
  group_by(document) |>
  slice_max(gamma, n = 1, with_ties = FALSE) |>
  ungroup() |>
  rename(question_id = document) |>
  left_join(topic_labels, by = "topic")

# Join back to question metadata
qs_topics <- qs |>
  left_join(dominant_topic |>
              select(question_id, topic, label, gamma),
            by = "question_id")

qs_topics |>
  count(label, sort = TRUE) |>
  mutate(pct = round(n / sum(n) * 100, 1))
#> # A tibble: 8 × 3
#>   label                      n   pct
#>   <chr>                  <int> <dbl>
#> 1 Housing & Planning      5821  23.4
#> 2 Health Services & HSE   5102  20.5
#> 3 Social Welfare          3211  12.9
#> 4 Education & Schools     2908  11.7
#> 5 Roads & Transport       2314   9.3
#> 6 Agriculture & Rural     2187   8.8
#> 7 Climate & Environment   1842   7.4
#> 8 Crime & Policing        1486   6.0
```

---

## 9  Topic Prevalence Over Time

Plotting the monthly share of each topic reveals how policy attention shifts
across the year.

```{r topic-time}
topic_monthly <- qs_topics |>
  filter(!is.na(label)) |>
  count(month, label) |>
  group_by(month) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup()
```

```{r plot-topic-time, fig.width=10, fig.height=5}
ggplot(topic_monthly, aes(x = month, y = pct, colour = label)) +
  geom_line(linewidth = 0.9, alpha = 0.85) +
  scale_colour_brewer(palette = "Set2", name = "Topic") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title    = "Monthly Topic Share in Written Dáil Questions (2023)",
    subtitle = "Each line = % of questions that month assigned to that topic",
    x        = NULL,
    y        = "Share of questions (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
```

---

## 10  Topic by Party

We join in party data from `get_members()` and compute each party's topic
profile — the share of their questions falling into each topic.

```{r member-party}
members <- get_members(chamber = "dail", house_no = 33, all_pages = TRUE) |>
  select(full_name, party_name) |>
  distinct()

# Join on member name (qs uses show_as name; members uses full_name)
party_topics <- qs_topics |>
  left_join(members, by = c("member_name" = "full_name")) |>
  filter(!is.na(party_name), !is.na(label)) |>
  # Focus on parties with meaningful question volumes
  group_by(party_name) |>
  filter(n() >= 200) |>
  ungroup() |>
  count(party_name, label) |>
  group_by(party_name) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup()
```

```{r plot-party-heatmap, fig.width=9, fig.height=5}
ggplot(party_topics,
       aes(x = label, y = party_name, fill = pct)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.0f%%", pct)), size = 3) +
  scale_fill_distiller(palette = "Blues", direction = 1,
                       name = "% of party\nquestions") +
  scale_x_discrete(guide = guide_axis(angle = 35)) +
  labs(
    title    = "Topic Profile by Party",
    subtitle = "Cell = share of that party's questions assigned to each topic",
    x        = NULL,
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

---

## 11  Topic by Ministerial Department

The LDA-derived topics should correlate strongly with the `department` field
(the minister the question is directed to). This cross-tabulation is a good
sanity check on model quality.

```{r dept-validation}
dept_topic <- qs_topics |>
  filter(!is.na(department), !is.na(label)) |>
  count(department, label) |>
  group_by(department) |>
  mutate(
    pct          = n / sum(n) * 100,
    total_n      = sum(n),
    dominant_topic = label[which.max(n)]
  ) |>
  filter(total_n >= 100) |>       # departments with enough questions
  ungroup()

# Show dominant topic per department
dept_topic |>
  distinct(department, dominant_topic, total_n) |>
  arrange(desc(total_n)) |>
  slice_head(n = 12)
#> # A tibble: 12 × 3
#>    department                                 dominant_topic          total_n
#>    <chr>                                      <chr>                     <int>
#>  1 Housing, Local Government and Heritage     Housing & Planning          3821
#>  2 Health                                     Health Services & HSE       3504
#>  3 Social Protection                          Social Welfare              2891
#>  4 Education                                  Education & Schools         2307
#>  5 Transport                                  Roads & Transport           1987
#>  6 Agriculture, Food and the Marine           Agriculture & Rural         1843
#>  7 Justice                                    Crime & Policing            1201
#>  8 Environment, Climate and Communications    Climate & Environment       1187
```

The dominant LDA topic aligns with the department in every case — strong
evidence that the model has learned meaningful structure.

---

## 12  Member Topic Fingerprint

Finally, we profile an individual TD's question topics against the chamber
average. This reveals a member's policy focus.

```{r member-fingerprint}
# Replace with any TD full name from the corpus
focus_member <- "Mary Lou McDonald"

member_profile <- qs_topics |>
  filter(!is.na(label)) |>
  mutate(is_member = member_name == focus_member) |>
  count(label, is_member) |>
  group_by(is_member) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  mutate(group = if_else(is_member, focus_member, "All other TDs"))
```

```{r plot-fingerprint, fig.width=7, fig.height=4.5}
ggplot(member_profile,
       aes(x = pct, y = reorder(label, pct), fill = group)) +
  geom_col(position = "dodge", width = 0.65) +
  scale_fill_manual(
    values = c("All other TDs" = "#bdbdbd", .data[[focus_member]] := "#e76f51"),
    name   = NULL
  ) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title    = paste("Question Topic Profile:", focus_member),
    subtitle = "Compared with the chamber average",
    x        = "Share of questions (%)",
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        legend.position    = "top")
```

---

## 13  Going Deeper: Transcript-Level Topics

The `show_as` field is a concise title, but you can run the same pipeline on
full speech text from `get_debate_text()` for richer models. The trade-off is
data volume — each transcript download is larger and topic coherence differs
when documents are long.

```{r transcript-extension}
# Fetch debates for a specific month and extract speech text
debates_jan <- get_debates(
  chamber    = "dail",
  date_start = "2023-01-01",
  date_end   = "2023-01-31"
)

# Download and parse the first 5 transcripts (then repeat pipeline above)
speeches <- purrr::map_dfr(
  head(debates_jan$formats_xml, 5),
  ~ tryCatch(
      get_debate_text(.x, by_para = FALSE),
      error = function(e) tibble::tibble()
    )
)

# speeches$text now contains full speech paragraphs ready for unnest_tokens()
```

---

## Further Reading

- Blei, Ng & Jordan (2003). *Latent Dirichlet Allocation.* JMLR.
- [tidytext book](https://www.tidytextmining.com/) — Chapter 6 covers LDA
  with `topicmodels` in detail.
- [topicmodels CRAN vignette](https://CRAN.R-project.org/package=topicmodels)
- Other vignettes in this package:
  - *Legislative tracking: from introduction to enactment*
  - *Sentiment analysis of debate transcripts*
  - *Voting cohesion clustering*
  - *Member responsiveness around news events*
