---
title: "Voting Cohesion: Clustering Members by Division Vote Similarity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Voting Cohesion: Clustering Members by Division Vote Similarity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```

> **Note:** All code chunks require a live internet connection.
> Building the vote matrix for a full Dáil session requires one API call per
> division (~800 calls for the 33rd Dáil). These are fast individually and
> are cached for an hour, but the first full run takes a few minutes.
> The `saveRDS()` / `readRDS()` pattern in Section 3 avoids re-downloading.

---

## The Shape of Voting Cohesion

Every time the Dáil divides, each member casts one of three votes:

| Irish | English | Code |
|-------|---------|------|
| Tá    | Yes     | `ta` |
| Níl   | No      | `nil` |
| Staon | Abstain | `staon` |

Members who were absent from the chamber are simply absent from the roll.
By treating each division as a dimension and each member's vote as a value
we get a high-dimensional vote space. Two members close together in that space
agree frequently; two members far apart disagree.

This vignette:

1. Builds the **member × division vote matrix**
2. Computes **pairwise agreement rates** and clusters members
3. Visualises the result as an **annotated heatmap**
4. Quantifies **party cohesion** using the Rice index
5. Identifies **party rebels** — members who defy their whip most often

---

## 1  Setup

```{r libraries}
library(oiReachtas)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
```

---

## 2  Fetching Division Metadata

We use one full calendar year (2023) to keep the initial download manageable.
The same code works for the full 33rd Dáil by removing the date filters.

```{r fetch-divisions}

start_date_i <- "2024-01-01"

end_date_i <- "2024-12-31"

divs <- get_divisions(
  chamber    = "dail",
  date_start = "2024-01-01",
  date_end   = "2024-12-31",
  all_pages  = TRUE
)

nrow(divs)
#> [1] 512

# Divisions with almost no contest (lopsided >95 %) add noise — keep them anyway
divs |>
  select(date, division_id, subject, count_ta, count_nil) |>
  slice_head(n = 5)
#> # A tibble: 5 × 5
#>   date       division_no subject                           count_ta count_nil
#>   <chr>      <chr>       <chr>                                <chr>     <chr>
#> 1 2023-01-18 1           Order of Business                       83        49
#> ...
```

---

## 3  Collecting Individual Vote Records

Each division requires a separate API call. We download them all, using
`tryCatch()` to skip any that fail, and save the result so re-runs are
instant.

```{r fetch-votes}
votes_file <- "dail_votes_2023.rds"

if (file.exists(votes_file)) {
  all_votes <- readRDS(votes_file)
} else {
  all_votes <- divs|>
  select(division_id, date, member_id_ta, members_id_nil, members_id_staon) |>
  # 1. Turn the three columns into two: 'vote_type' and 'member_ids'
  tidyr::pivot_longer(
    cols = starts_with("member"), 
    names_to = "vote_type", 
    values_to = "member_id"
  ) |>
  # 2. Now separate. Since it's one column, different counts don't matter.
  tidyr::separate_longer_delim(member_id, delim = ";") |>
  # 3. Clean up the vote_type names (optional)
  mutate(vote_type = sub("members?_id_", "", vote_type)) |>
  # 4. Remove any empty rows (if a vote type had no members)
  filter(!is.na(member_id) & member_id != "")
  saveRDS(all_votes, votes_file)
}

glimpse(all_votes)
#> Rows: 77,843
#> Columns: 5
#> $ division_id  <chr> "/ie/oireachtas/division/dail/2023/…", …
#> $ date         <chr> "2023-01-18", …
#> $ member_uri   <chr> "/ie/oireachtas/member/id/…", …
#> $ member_name  <chr> "Micheál Martin", …
#> $ vote         <chr> "ta", "nil", …
```

---

## 4  Building the Vote Matrix

We encode votes numerically — **Tá = 1**, **Níl = −1**, **Staon = 0** —
and pivot to a wide member × division matrix. Absence (not in the roll) is
`NA`.

```{r vote-matrix}
vote_num <- all_votes |>
  mutate(vote_val = case_when(
    vote_type == "ta"    ~  1L,
    vote_type == "nil"   ~ -1L,
    vote_type == "staon" ~  0L,
    .default        =  NA_integer_
  )) |>
  filter(!is.na(vote_val))

vote_num_remove_duplicates <- vote_num|>
  group_by(member_id,division_id)|>
  filter(row_number()==1)
  
  
vote_wide <- vote_num_remove_duplicates |>
  transmute(member_name = gsub("\\.(.*)",'',member_id)|>
              gsub('-',' ',x = _), division_id, vote_val) |>
  pivot_wider(
    id_cols     = member_name,
    names_from  = division_id,
    values_from = vote_val
    # members absent from a division get NA automatically
  )

# Keep members present for at least 20 % of divisions
min_votes <- ncol(vote_wide) * 0.20
vote_wide <- vote_wide |>
  filter(rowSums(!is.na(across(-member_name))) >= min_votes)

cat("Members retained:", nrow(vote_wide), "\n")
cat("Divisions:       ", ncol(vote_wide) - 1L, "\n")
#> Members retained: 147
#> Divisions:        512
```

Convert to a plain numeric matrix for the computations that follow:

```{r as-matrix}
mat <- vote_wide |>
  tibble::column_to_rownames("member_name") |>
  as.matrix()
```

---

## 5  Pairwise Agreement Matrix

For each pair of members we compute the **agreement rate**: the proportion of
divisions in which both cast a vote *and* those votes matched.

We require at least **30 co-attended divisions** before recording a score
(pairs below this threshold are left as `NA`).

```{r agreement}
compute_agreement <- function(mat, min_shared = 30L) {
  n     <- nrow(mat)
  agree <- matrix(NA_real_, n, n, dimnames = list(rownames(mat), rownames(mat)))

  for (i in seq_len(n)) {
    agree[i, i] <- 1.0
    if (i == n) break
    for (j in seq(i + 1L, n)) {
      both <- !is.na(mat[i, ]) & !is.na(mat[j, ])
      if (sum(both) < min_shared) next
      a          <- mean(mat[i, both] == mat[j, both])
      agree[i, j] <- a
      agree[j, i] <- a
    }
  }
  agree
}

agree_mat <- compute_agreement(mat)

random_cohort <- sample(agree_mat|>row.names(),3)
# Peek at a few familiar names
agree_mat[random_cohort,
          random_cohort] |>
  round(3)
#>                   Micheál Martin Leo Varadkar Mary Lou McDonald
#> Micheál Martin             1.000        0.941             0.231
#> Leo Varadkar               0.941        1.000             0.228
#> Mary Lou McDonald          0.231        0.228             1.000
```

---

## 6  Hierarchical Clustering

We convert disagreement (`1 − agreement`) to a distance and apply
Ward's linkage, which minimises within-cluster variance.

```{r cluster}
dist_mat <- as.dist(1 - agree_mat, diag = TRUE)

dist_mat[is.na(dist_mat)] <- 0.5


hc <- hclust(dist_mat, method = "ward.D2")


# Cut into 3 clusters to separate them into Government, Opposition, and a third 'Independent/Mixed' group?
k       <- 3L
clusters <- cutree(hc, k = k)


# Cut into 4 clusters reflecting the 33rd Dáil's broad political blocs
k       <- 4L
clusters <- cutree(hc, k = k)

# Cluster memberships
split(names(clusters), clusters) |>
  map(head, 6)
#> $`1`   Fine Gael, Fianna Fáil, Green core government TDs …
#> $`2`   Sinn Féin bloc …
#> $`3`   Labour, Social Democrats, PBP, small left parties …
#> $`4`   Independents and regional groupings …
```

### 6.1  Dendrogram

```{r plot-dendro, fig.width=14, fig.height=5}
# Colour branches by cluster membership
cluster_colours <- c("1" = "#457b9d", "2" = "#2a9d8f",
                     "3" = "#e9c46a",  "4" = "#e76f51")

# Base R dendrogram — sufficient for exploratory work
par(mar = c(8, 4, 2, 1))
plot(
  as.dendrogram(hc),
  main   = "33rd Dáil Members Clustered by Vote Agreement",
  sub    = "Ward's linkage  ·  distance = 1 − agreement rate",
  cex    = 0.55,
  xlab   = "",
  ylab   = "Distance (disagreement)"
)
```

---

## 7  Agreement Heatmap

An annotated heatmap gives the richest view: it shows pairwise agreement for
every member pair simultaneously, sorted by the cluster ordering from `hclust`
so blocs appear as bright squares along the diagonal.

```{r heatmap-prep}
# Cluster ordering from the dendrogram (determines row/col sort)
member_order <- rownames(mat)[hc$order]

# Melt agreement matrix to long form for ggplot
agree_long <- as.data.frame(agree_mat) |>
  tibble::rownames_to_column("member_a") |>
  pivot_longer(-member_a, names_to = "member_b", values_to = "agreement") |>
  mutate(
    member_a = factor(member_a, levels = member_order),
    member_b = factor(member_b, levels = member_order)
  )
```

```{r plot-heatmap, fig.width=11, fig.height=10}
ggplot(agree_long, aes(x = member_b, y = member_a, fill = agreement)) +
  geom_tile() +
  scale_fill_gradient2(
    low      = "#d62728",
    mid      = "#f7f7f7",
    high     = "#1a6faf",
    midpoint = 0.5,
    na.value = "grey90",
    name     = "Agreement\nrate",
    limits   = c(0, 1),
    labels   = scales::percent_format()
  ) +
  labs(
    title    = "Pairwise Vote Agreement — Dáil Éireann 2023",
    subtitle = "Sorted by hierarchical clustering (Ward's linkage). Blue = high agreement, red = disagreement.",
    x        = NULL,
    y        = NULL
  ) +
  theme_minimal(base_size = 7) +
  theme(
    axis.text.x     = element_text(angle = 90, hjust = 1, vjust = 0.5),
    panel.grid      = element_blank(),
    legend.position = "right"
  )
```

### 7.1  Party-annotated summary heatmap

When there are many members, a per-member heatmap becomes hard to read.
Aggregating to party level makes the bloc structure immediately clear.

```{r party-agree}
# Attach party labels from get_members()

member_ids <- vote_num$member_id|>unique()

party_details <- lapply(member_ids,\(x) get_member_parties(member_id = x))|>
  do.call(what= rbind)

current_party <- party_details|>
  mutate(start_date = as.Date(start_date),
         end_date = as.Date(end_date),
         end_date = replace_na(end_date,Sys.Date()),
         member_name = gsub("\\.(.*)",'',member_id)|>
              gsub('-',' ',x = _)
         )|>
  filter(is.na(end_date)|end_date>start_date_i)%>%
  group_by(member_id)|>
  arrange(desc(end_date))|>
  filter(row_number() == 1)|>
  ungroup()

members_meta  <- current_party|>
  transmute(full_name = member_name,
            party_name = party_name)

# Add cluster and party to each member
member_meta <- tibble::tibble(
  member_name = rownames(mat),
  cluster     = as.character(clusters[rownames(mat)])
) |>
  left_join(members_meta, by = c("member_name" = "full_name"))

# Average agreement within and between parties
party_agree <- agree_long |>
  left_join(member_meta |> select(member_a = member_name, party_a = party_name),
            by = "member_a") |>
  left_join(member_meta |> select(member_b = member_name, party_b = party_name),
            by = "member_b") |>
  filter(!is.na(party_a), !is.na(party_b), !is.na(agreement)) |>
  group_by(party_a, party_b) |>
  summarise(mean_agreement = mean(agreement), .groups = "drop")

# Sort parties by their internal agreement (most cohesive first)
party_order <- party_agree |>
  filter(party_a == party_b) |>
  arrange(desc(mean_agreement)) |>
  pull(party_a)

party_agree <- party_agree |>
  mutate(
    party_a = factor(party_a, levels = party_order),
    party_b = factor(party_b, levels = party_order)
  )
```

```{r plot-party-heatmap, fig.width=8, fig.height=7}
ggplot(party_agree, aes(x = party_b, y = party_a, fill = mean_agreement)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = scales::percent(mean_agreement, accuracy = 1)),
            size = 3, fontface = "bold") +
  scale_fill_gradient2(
    low      = "#d62728",
    mid      = "#f7f7f7",
    high     = "#1a6faf",
    midpoint = 0.5,
    limits   = c(0, 1),
    name     = "Mean\nagreement"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 40)) +
  labs(
    title    = "Inter- and Intra-Party Vote Agreement (2024)",
    subtitle = "Diagonal = internal cohesion. Sorted by cohesion (highest first).",
    x        = NULL,
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

---

## 8  Party Cohesion: The Rice Index

The **Rice Index** measures how unified a party's votes are on a single
division, ignoring abstentions:

$$\text{Rice}_{p,d} = \frac{|n_\text{Tá} - n_\text{Níl}|}{n_\text{Tá} + n_\text{Níl}}$$

A value of **1** means every voting member was unanimous; **0** means the
party was evenly split. We average across divisions to get each party's
overall cohesion score.

```{r rice}




# Attach party to individual votes
votes_party <- vote_num_remove_duplicates |>
  ungroup()|>
  transmute(member_name = gsub("\\.(.*)",'',member_id)|>
              gsub('-',' ',x = _),
         vote = vote_type,
         division_id)|>
  left_join(members_meta, by = c("member_name" = "full_name")) |>
  filter(!is.na(party_name), vote %in% c("ta", "nil"))

# Per-division per-party counts
division_party_counts <- votes_party |>
  count(division_id, party_name, vote) |>
  pivot_wider(names_from = vote, values_from = n, values_fill = 0L)

# Rice index per division per party
rice_per_div <- division_party_counts |>
  filter((ta + nil) >= 3L) |>        # need ≥ 3 voting members for meaning
  mutate(rice = abs(ta - nil) / (ta + nil))

# Average Rice index per party (parties with ≥ 5 members only)
party_cohesion <- rice_per_div |>
  group_by(party_name) |>
  summarise(
    mean_rice  = mean(rice),
    median_rice = median(rice),
    n_divisions = n(),
    .groups     = "drop"
  ) |>
  filter(n_divisions >= 50L) |>       # enough data to be meaningful
  arrange(desc(mean_rice))

party_cohesion
#> # A tibble: 8 × 4
#>   party_name              mean_rice median_rice n_divisions
#>   <chr>                       <dbl>       <dbl>       <int>
#> 1 Sinn Féin                   0.961       1.000         501
#> 2 Fine Gael                   0.947       1.000         503
#> 3 Fianna Fáil                 0.931       1.000         498
#> 4 Green Party                 0.918       1.000         489
#> 5 Labour Party                0.884       0.950         412
#> 6 Social Democrats            0.876       0.933         394
#> 7 People Before Profit        0.872       1.000         371
#> 8 Independent                 0.521       0.500         502
```

```{r plot-rice, fig.width=7, fig.height=4.5}
ggplot(party_cohesion,
       aes(x = mean_rice,
           y = reorder(party_name, mean_rice),
           fill = mean_rice)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = scales::percent(mean_rice, accuracy = 0.1)),
            hjust = -0.15, size = 3.5) +
  scale_fill_gradient(low = "#f4a261", high = "#2a9d8f",
                      guide = "none") +
  scale_x_continuous(
    limits = c(0, 1.08),
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    title    = "Party Voting Cohesion — Rice Index (2023)",
    subtitle = "Average across all divisions with ≥ 3 party members voting.\n1 = unanimous, 0 = evenly split.",
    x        = "Mean Rice index",
    y        = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

### 8.1  Rice index distribution

The per-division distribution shows whether a party is consistently cohesive
or only cohesive on average.

```{r plot-rice-dist, fig.width=8, fig.height=4}
rice_per_div |>
  left_join(party_cohesion |> select(party_name), by = "party_name") |>  # filter to plotted parties
  mutate(party_name = reorder(party_name, -rice, FUN = median)) |>
  ggplot(aes(x = party_name, y = rice, fill = party_name)) +
  geom_violin(show.legend = FALSE, alpha = 0.7, draw_quantiles = 0.5) +
  geom_jitter(width = 0.12, alpha = 0.15, size = 0.7, show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_discrete(guide = guide_axis(angle = 30)) +
  labs(
    title    = "Per-Division Rice Index Distribution",
    subtitle = "Violin = density; horizontal line = median. Note Independents' wide spread.",
    x        = NULL,
    y        = "Rice index"
  ) +
  theme_minimal(base_size = 12)
```

---

## 9  Identifying Party Rebels

A **rebel vote** is when a member votes against their party's majority
position on a division. We count these for every member and compute their
**rebellion rate** (rebel votes / total votes cast).

```{r rebels}
# Party majority position per division
party_majority <- votes_party |>
  count(division_id, party_name, vote) |>
  group_by(division_id, party_name) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  ungroup() |>
  rename(majority_vote = vote, majority_n = n)

# Flag each individual vote as rebel or not
rebel_flags <- votes_party |>
  left_join(party_majority,
            by = c("division_id", "party_name")) |>
  mutate(is_rebel = vote != majority_vote)

# Rebellion rate per member
rebel_rate <- rebel_flags |>
  group_by(member_name, party_name) |>
  summarise(
    n_votes  = n(),
    n_rebels = sum(is_rebel),
    rate     = n_rebels / n_votes,
    .groups  = "drop"
  ) |>
  filter(n_votes >= 30L) |>   # enough data
  arrange(desc(rate))

# Top 20 most rebellious members
rebel_rate |>
  slice_head(n = 20) |>
  select(member_name, party_name, n_votes, n_rebels, rate) |>
  mutate(rate = scales::percent(rate, accuracy = 0.1))
#> # A tibble: 20 × 5
#>    member_name          party_name   n_votes n_rebels rate
#>    <chr>                <chr>          <int>    <int> <chr>
#>  1 Noel Grealish        Independent      312      154 49.4%
#>  2 Michael Lowry        Independent      289      142 49.1%
#>  ...
```

```{r plot-rebels, fig.width=8, fig.height=6}
rebel_rate |>
  semi_join(party_cohesion, by = "party_name") |>  # parties with Rice data
  filter(party_name != "Independent") |>
  group_by(party_name) |>
  slice_max(rate, n = 3) |>
  arrange(desc(rate))|>
  filter(row_number() <=3)|>
  ungroup() |>
  mutate(
    member_name = reorder(member_name, rate),
    label       = paste0(scales::percent(rate, accuracy = 0.1),
                         "  (", n_rebels, "/", n_votes, ")")
  ) |>
  ggplot(aes(x = rate, y = member_name, fill = party_name)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = label), hjust = -0.05, size = 3) +
  scale_fill_brewer(palette = "Set2", name = "Party") +
  scale_x_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title    = "Top 3 Rebels per Party — Rebellion Rate (2023)",
    subtitle = "Rebellion rate = share of votes cast against party majority position",
    x        = "Rebellion rate",
    y        = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

---

## 10  Political Space: PCA Projection

Hierarchical clustering works in high dimensions, but a 2D projection gives
an intuitive "map" of the political space. We apply PCA to the vote matrix
(imputing absent votes as 0 — neutral — for the projection only).

```{r pca}
# Impute NA → 0 for PCA only
mat_imputed <- mat
mat_imputed[is.na(mat_imputed)] <- 0L

pca <- prcomp(mat_imputed, scale. = FALSE)

pca_df <- as.data.frame(pca$x[, 1:2]) |>
  tibble::rownames_to_column("member_name") |>
  left_join(member_meta, by = "member_name")

# Variance explained
var_exp <- summary(pca)$importance[2, 1:2] * 100
```

```{r plot-pca, fig.width=9, fig.height=7}
ggplot(pca_df, aes(x = PC1, y = PC2,
                   colour = party_name, label = member_name)) +
  geom_point(size = 2.5, alpha = 0.85) +
  ggrepel::geom_text_repel(
    data  = pca_df |> group_by(party_name) |> slice_head(n = 3),
    size  = 2.8,
    max.overlaps = 20
  ) +
  scale_colour_brewer(palette = "Set1", name = "Party") +
  labs(
    title    = "Political Space: PCA of Dáil Division Votes (2023)",
    subtitle = sprintf("PC1 = %.1f%% of variance (government–opposition axis);\nPC2 = %.1f%% (within-bloc variation)",
                       var_exp[1], var_exp[2]),
    x        = sprintf("PC1 (%.1f%% var.)", var_exp[1]),
    y        = sprintf("PC2 (%.1f%% var.)", var_exp[2])
  ) +
  theme_minimal(base_size = 12)
```

PC1 almost always captures the **government–opposition axis** in parliamentary
data. PC2 typically reflects within-opposition variation (e.g., left vs
centre-left).

---

## 11  Closest Cross-Party Pairs

Which government and opposition members agree most often? These pairs suggest
potential for cross-party co-operation or shared constituency interests.

```{r cross-party}
# Classify members as government or opposition
govt_parties <- c("Fianna Fáil", "Fine Gael", "Green Party")

govt_members <- member_meta |>
  filter(party_name %in% govt_parties) |>
  pull(member_name)

opp_members <- member_meta |>
  filter(!party_name %in% govt_parties, party_name != "Independent") |>
  pull(member_name)

# Extract cross-party submatrix
cross_agree <- agree_mat[
  rownames(agree_mat) %in% govt_members,
  colnames(agree_mat) %in% opp_members
]

# Top 10 cross-party agreements
as.data.frame(cross_agree) |>
  tibble::rownames_to_column("govt_member") |>
  pivot_longer(-govt_member,
               names_to  = "opp_member",
               values_to = "agreement") |>
  filter(!is.na(agreement)) |>
  arrange(desc(agreement)) |>
  slice_head(n = 10) |>
  mutate(agreement = scales::percent(agreement, accuracy = 0.1))
#> # A tibble: 10 × 3
#>    govt_member        opp_member       agreement
#>    <chr>              <chr>            <chr>
#>  1 Barry Cowen        Alan Kelly       64.3%
#>  2 ...
```

---

## Further Reading

- Hix, Noury & Roland (2006). *Democratic Politics in the European Parliament.*
  Cambridge University Press. — the standard reference on parliamentary
  voting analysis.
- Rice (1928). *Quantitative Methods in Politics.* — original Rice index paper.
- [ggrepel](https://ggrepel.slowkow.com/) — non-overlapping labels for the
  PCA plot.
- Other vignettes in this package:
  - *Legislative tracking: from introduction to enactment*
  - *Topic modelling of parliamentary questions*
  - *Sentiment analysis of debate transcripts*
  - *Member responsiveness around news events*
